# Individual Geoms

```{r 03-individualgeoms, include=FALSE}

library(ggplot2)
library(tidyverse)
```

- Geoms are the fundamental building blocks of ggplot2.
- Most of the geoms are associated with a named plot.
- Some geoms can be added on to low-level geoms to create more complex plots.
- To find out more about individual geoms see their documentation.

## Scatterplot:
```{r 03-scatter}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()
```

##  Line plot:
```{r 03-lineplot}
ggplot(economics, aes(date, unemploy / pop)) +
  geom_line() 
```

## Histogram: 
```{r 03-hist}
ggplot(mpg, aes(hwy)) + geom_histogram() 
```

## Bar chart
```{r 03-bar}
ggplot(mpg, aes(manufacturer)) + 
  geom_bar() 
```
## geom_path() connects points in order of appearance.
```{r 03-df, include=FALSE}
df <- data.frame(
  x = c(3, 1, 5), 
  y = c(2, 4, 6), 
  label = c("a","b","c")
)
p <- ggplot(df, aes(x, y, label = label)) + 
  labs(x = NULL, y = NULL) + # Hide axis label
  theme(plot.title = element_text(size = 12))
```

```{r 03-path}
p + geom_path()
```

## geom_polygon() draws polygons which are filled paths.
```{r 03-poly}
p + geom_polygon() 
```

## geom_line() connects points from left to right.
```{r 03-line}
p + geom_line() 
```


## What low-level geoms are used to draw geom_smooth()? 
Geom_smooth() fits a smoother to data, displaying the smooth and its standard error, allowing you to see a dominant pattern within a scatterplot with a lot of "noise". The low level geom for geom_smooth() are geom_path(), geom_area() and geom_point().

```{r 03-smooth}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth()
```

## What low-level geoms are used to draw geom_boxplot()? 
Box plots are used to summarize the distribution of a set of points using summary statistics. The low level geom for geom_boxplot() are geom_rect(), geom_line() and geom_point(). 

```{r 03-box}
ggplot(mpg, aes(drv, hwy)) + geom_boxplot()
```

## What low-level geoms are used to draw geom_violin()?
Violin plots show a compact representation of the density of the distribution highlighting the areas where most of the points are found. The low level geom for geom_violin() are geom_area() and geom_path(). 

```{r 03-violin}
ggplot(mpg, aes(drv, hwy)) + geom_violin()
```


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/XcTYBH4XbHo")`

<details>
  <summary> Meeting chat log </summary>
```
00:13:39	priyanka gagneja:	I forgot to mention that since this is relatively smaller chapter Ryan has prepared some material introducing Chapter 4 for today and he will talking about the entire chapter next week.
00:16:38	priyanka gagneja:	that's correct
00:16:42	priyanka gagneja:	that's my understanding too
00:18:38	priyanka gagneja:	what do you mean circles .. can you share a more detailed example
00:21:59	Jiwan Heo:	tibble(id = 1:10) %>% mutate(x = cos(2*pi*id/10), y = sin(2*pi*id/10)) %>% ggplot(aes(x, y)) + geom_line() + coord_equal()
00:22:05	Jiwan Heo:	vs tibble(id = 1:10) %>% mutate(x = cos(2*pi*id/10), y = sin(2*pi*id/10)) %>% ggplot(aes(x, y)) + geom_path() + coord_equal()
00:35:42	priyanka gagneja:	Thank you Ryan !!
00:38:16	priyanka gagneja:	need a min
00:52:22	Michael Haugen:	“Side rail” no pun intended
00:52:34	Ryan S:	lol
00:52:34	Michael Haugen:	sounds great
00:54:02	Ryan Metcalf:	I was going to use Derail…..no pun intended!
00:54:05	Ryan Metcalf:	Thanks you everyone!
```
</details>
