---
output:
  pdf_document: default
  html_document: default
---

# Layers

*Introduction*

- Based on the *Grammar of Graphics*, GGplot2 encourages you to build plots in a structured manner building upon on your plots with various layers. 
- The purposes of the layers are: 
- to display the **data**. 
- to display a **statistical summary** of the data.
- to add **metadata** such as context, annotations or references to your plot.

## Layers: Chapters

**The section on layers is broken up into the following chapters:**

- Chapter 3. Individual Geoms
- Chapter 4. Collective Geoms
- Chapter 5. Statistical Summaries
- Chapter 6. Maps
- Chapter 7. Networks
- Chapter 8. Annotations
- Chapter 9. Arranging Plots

## Chapter 3: Individual Geoms
```{r include=FALSE}

library(ggplot2)
library(tidyverse)
```
- Geoms are the fundamental building blocks of ggplot2.
- Most of the geoms are associated with a named plot.
- Some geoms can be added on to low-level geoms to create more complex plots.
- To find out more about individual geoms see their documentation.

## Scatterplot:
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()
```

##  Line plot:
```{r}
ggplot(economics, aes(date, unemploy / pop)) +
  geom_line() 
```

## Histogram: 
```{r}
ggplot(mpg, aes(hwy)) + geom_histogram() 
```

## Bar chart
```{r}
ggplot(mpg, aes(manufacturer)) + 
  geom_bar() 
```
## geom_path() connects points in order of appearance.
```{r include=FALSE}
df <- data.frame(
  x = c(3, 1, 5), 
  y = c(2, 4, 6), 
  label = c("a","b","c")
)
p <- ggplot(df, aes(x, y, label = label)) + 
  labs(x = NULL, y = NULL) + # Hide axis label
  theme(plot.title = element_text(size = 12))
```

```{r}
p + geom_path()
```

## geom_polygon() draws polygons which are filled paths.
```{r}
p + geom_polygon() 
```

## geom_line() connects points from left to right.
```{r}
p + geom_line() 
```


## What low-level geoms are used to draw geom_smooth()? 
Geom_smooth() fits a smoother to data, displaying the smooth and its standard error, allowing you to see a dominant pattern within a scatterplot with a lot of "noise". The low level geom for geom_smooth() are geom_path(), geom_area() and geom_line().

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth()
```

## What low-level geoms are used to draw geom_boxplot()? 
Box plots are used to summarize the distribution of a set of points using summary statistics. The low level geom for geom_boxplot() are geom_rect(), geom_line() and geom_point(). 

```{r}
ggplot(mpg, aes(drv, hwy)) + geom_boxplot()
```

## What low-level geoms are used to draw geom_violin()?
Violin plots show a compact representation of the density of the distribution highlighting the areas where most of the points are found. The low level geom for geom_violin() are geom_area() and geom_path(). 

```{r}
ggplot(mpg, aes(drv, hwy)) + geom_violin()
```

