# Maps

**Learning Objectives:**
 - Plot simple maps using `geom_polygon()`
 - Using *simple features* `sf` to plot GIS data `geom_sf()`
 - Work with map projections and underlying `sf` data structure
 - Draw maps using Raster data
 
Plotting geospacial data is a common visualization task. The process may require spcialized tools. You can decompse the problem into two paths:
  - Using one data source to draw a map (if you have GIS data)
  - Adding metadata from another information source to the map (more common with relation to geographic areas)
  
> NOTE: X = Longitude, Y=Latitude. When pronounced "Lat/Lon" it is actually measured as Y/X. Not confusing....just keeping with vocabulary and measurements!

## Polygon Maps
The simplest approach to mapping is using geom_polygon(). This forms bounderies around regions.

```{r 06-mi-countries, include=TRUE}
library(ggplot2)
mi_counties <- map_data("county", "michigan") %>% 
  select(lon = long, lat, group, id = subregion)
head(mi_counties)
```

In this data set we have four variables:
 - `lat`: Latitude of the vertex (as measured by horizontal paths)
 - `long`: Longitude of the vertex (as measured by vertical paths)
 - `id`: name of the region
 - `group`: unique identifier for contiguous areas within a region
 
```{r 06-coord_quickmap(), include=TRUE}
ggplot(mi_counties, aes(lon, lat)) + 
  geom_point(size = .25, show.legend = FALSE) +
  coord_quickmap()
ggplot(mi_counties, aes(lon, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap()
```

In this plot, `coord_quickmap()` is used to adjust the axes to ensure longitude and latitude are rendered on the same scale

For a more advanced use of ggplot2 for mapping, we'll see the use of `geom_sf()` and `coord_sf()` to handle spatial data specified in simple features format.

## Simple Features Maps
You *can* use the above examples...but not real world pratical. Instead, most GIS data is written as *simple features* and produced by the (Open Geospatial Consortium]https://www.ogc.org/)

### Layered Maps
### Labelled Maps
### Adding Other Geoms
## Map Projections
## Working with `sf` Data
## Raster Maps
## Data Sources


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/8-_iLnAogss")`

<details>
  <summary> Meeting chat log </summary>
```
00:11:25	June Choe:	hello!
00:15:21	SriRam:	Hi all, I am new here, I came to know about this from ISLR book club
00:16:14	Stan Piotrowski:	Great have to have you here, SriRam!  Some of us are also in the ISLR book club and I think this is a nice complement to that material
00:25:26	June Choe:	I'd like to see the error!
00:26:54	June Choe:	I think you'd have to add a geom_labe() layer
00:27:08	June Choe:	but as Stan said it'll render text at every point
00:27:29	June Choe:	after polygon would draw it on top
00:33:36	Michael Haugen:	Reminds me of a Flight of the Concords episode
00:35:04	SriRam:	23.5
00:40:09	SriRam:	It would be incorrect data to have multiple geometries on same record
00:47:44	Lydia Gibson:	It’s spelled right… or at least that’s how it’s spelled in the book
00:48:16	June Choe:	hm maybe the sf_label and sf_text layers also need to take the geometry aesthetic
00:49:02	June Choe:	label.padding I think is from geom_label (the white space between text and bounding box)
00:52:51	Federica Gazzelloni:	viridis
00:53:30	Federica Gazzelloni:	scale_color_viridis()
00:53:42	Federica Gazzelloni:	scale_fill_viridis()
00:54:01	SriRam:	I think viridis is better for continuous values
00:54:57	Federica Gazzelloni:	viridian package: https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
00:55:08	Federica Gazzelloni:	viridis package: https://cran.r-project.org/web/packages/viridis/vignettes/intro-to-viridis.html
00:55:51	Federica Gazzelloni:	more: https://www.rdocumentation.org/packages/viridis/versions/0.5.1/topics/scale_color_viridis
00:56:21	Michael Haugen:	David Robinson uses scale_fill_viridis_c() for a map in his most recent Tidy Tuesday Screen cast. See around 23minute mark: Tidy Tuesday live screencast: Analyzing registered nurses in R. https://www.youtube.com/watch?v=UVmxHb2Daeo&t=486s
01:12:20	Lydia Gibson:	Thank you Ryan!!
01:12:34	Federica Gazzelloni:	thanks Ryan
01:12:54	Stan Piotrowski:	Thanks Ryan!
```
</details>

`r knitr::include_url("https://www.youtube.com/embed/RiACZPhm53Y")`

<details>
  <summary> Meeting chat log </summary>
```
00:09:20	priyanka gagneja:	sorry everyone I just joined
00:09:38	Federica Gazzelloni:	Hello!
00:10:02	priyanka gagneja:	and will probably be a little in and out .. got a not so happy baby today at home
00:18:11	Stan Piotrowski:	I need to take off for a conflict that just came up.  Catch up with you all on slack!
00:20:46	SriRam:	It is the image product ID
00:21:07	SriRam:	All the IDE’s
00:21:50	Kent Johnson:	The IDE codes are defined here: https://ropensci.github.io/bomrang/reference/get_available_imagery.html
00:27:32	SriRam:	The process is called geo-referencing
00:27:58	SriRam:	And image is called a geo-referenced image
00:31:33	SriRam:	Yes, it is a reference system
00:31:38	SriRam:	A coordinate reference
00:33:04	Federica Gazzelloni:	this is the bit that makes the reference: crs = st_crs(sat_vis)
00:49:27	Jiwan Heo:	something just came up, and have to leave. See you all next week!
00:59:58	priyanka gagneja:	I am signing off now , can someone please address and sign off on my behalf. I will send a msg later on slack
```
</details>
