# Maps

**Learning Objectives:**
 - Plot simple maps using `geom_polygon()`
 - Using *simple features* `sf` to plot GIS data `geom_sf()`
 - Work with map projections and underlying `sf` data structure
 - Draw maps using Raster data
 
Plotting geospacial data is a common visualization task. The process may require spcialized tools. You can decompse the problem into two paths:
  - Using one data source to draw a map (if you have GIS data)
  - Adding metadata from another information source to the map (more common with relation to geographic areas)
  
> NOTE: X = Longitude, Y=Latitude. When pronounced "Lat/Lon" it is actually measured as Y/X. Not confusing....just keeping with vocabulary and measurements!

## Polygon Maps
The simplest approach to mapping is using geom_polygon(). This forms bounderies around regions.

```{r 06-mi-countries, include=TRUE}
library(ggplot2)
mi_counties <- map_data("county", "michigan") %>% 
  select(lon = long, lat, group, id = subregion)
head(mi_counties)
```

In this data set we have four variables:
 - `lat`: Latitude of the vertex (as measured by horizontal paths)
 - `long`: Longitude of the vertex (as measured by vertical paths)
 - `id`: name of the region
 - `group`: unique identifier for contiguous areas within a region
 
```{r 06-coord_quickmap(), include=TRUE}
ggplot(mi_counties, aes(lon, lat)) + 
  geom_point(size = .25, show.legend = FALSE) +
  coord_quickmap()
ggplot(mi_counties, aes(lon, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap()
```

In this plot, `coord_quickmap()` is used to adjust the axes to ensure longitude and latitude are rendered on the same scale

For a more advanced use of ggplot2 for mapping, we'll see the use of `geom_sf()` and `coord_sf()` to handle spatial data specified in simple features format.

## Simple Features Maps
You *can* use the above examples...but not real world pratical. Instead, most GIS data is written as *simple features* and produced by the (Open Geospatial Consortium]https://www.ogc.org/)

### Layered Maps
### Labelled Maps
### Adding Other Geoms
## Map Projections
## Working with `sf` Data
## Raster Maps
## Data Sources