# Networks

**Learning Objectives**

- What is network data?
- New functions and geoms
- Visualization of nodes and edges as abstract concepts 

## Introduction {-}

This chapter illustrates how to visualize network data using additional packages that go beyond what `{ggplot2}` is able to do natively.

- `{tidygraph}` tidy graph manipulation
- `{ggraph}`    network visualization
- `{igraph}`    generating random and regular graphs (and viz in base R)

## What is network data? {-}

```{r echo = FALSE, warning = FALSE, message = FALSE}
set.seed(5)
library(tidygraph)
library(ggraph)
g <- tidygraph::play_gnp(n = 10, p = 0.5, directed = F)
ggraph(g)+
  geom_edge_link() +
  geom_node_point(col = "dodgerblue", size = 6)+
  theme_graph()
```

Network data consists of entities (**nodes** or vertices) and their relations (**edges** or links).

Edges: directed/undirected, weighted/unweighted

Examples:

* Social network: friendships (edges) between people (nodes)
* A food web: trophic relationships (edges) between species (nodes)
* Plant-pollinator networks

## Network data is special {-}

- Must represent both **nodes** and **edges**
- Two main ways of representing network data:
  1. Edge list (long format)
```{r echo = F, message = F}
set.seed(3)
edge_list <- 
  data.frame(from = c(letters[1:4], letters[1:3]),
         to = c(letters[1:4], letters[2:4]),
         weight = c(rep(1, 4), runif(3)))
edge_list
```
  
  2. Adjacency matrix (wide format)
```{r echo = F}
g <- igraph::as_adjacency_matrix(igraph::simplify(igraph::graph_from_data_frame(edge_list)))
g
```



## {tidygraph}: A tidy network manipulation API {-}

- A {dplyr} API for network data

New functions:

- `activate()` tells tidygraph which part of the network you want to focus on, either nodes or edges
- `.N()` which gives access to the node data of the current graph even when working with the edges
- `.E()` and `.G()` to access the edges or the whole graph)

```{r 07-01, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
```

## Example: creating a graph {-}

In this **example** we create a graph, assign a random label to the nodes, and sort the edges based on the label of their source node.

The function `play_gnp()` creates graphs directly through sampling of different attributes.

```{r 07-play_gnp, message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidygraph)

graph <- tidygraph::play_gnp(n = 10, p = 0.2) %>% 
  activate(nodes) %>% 
  mutate(class = sample(letters[1:4], n(), replace = TRUE)) %>% 
  activate(edges) %>% 
  arrange(.N()$class[from])

graph
```



### Conversion to `tbl_graph` {-}

Convert data with `as_tbl_graph()`. A `tbl_graph` is a data structure for tidy graph manipulation. It converts a data frame encoded as an edgelist.

```{r 07-highschool}
data(highschool, package = "ggraph")
head(highschool)
```

With `as_tbl_graph()` we get:

```{r 07-hs_graph}
hs_graph <- tidygraph::as_tbl_graph(highschool, directed = FALSE)
hs_graph
```

## Example: colors {-}

Other data that's in a network format:

In this **example** the `luv_colours()` function allows for all built-in `colors()` translated into **Luv colour space**, a data frame with 657 observations and 4 variables:
[luv_colours](https://github.com/tidyverse/ggplot2/blob/main/data-raw/luv_colours.R)

```{r 07-luv_colours}
luv_colours <- as.data.frame(convertColor(t(col2rgb(colors())),
                                          "sRGB", "Luv"))
luv_colours$col <- colors()
head(luv_colours)
```

## {-}

This visualization represents the content of the dataset. Then we will see how it looks in a graph representation. 

```{r 07-colors}
ggplot(luv_colours, aes(u, v)) + 
geom_point(aes(colour = col), size = 3) + 
scale_color_identity() + 
coord_equal() +
  theme_void()
```

## {-}

We notice some colors are closer to each other than others. We might want to use a clustering algorithm to see how they relate to each other.

```{r 07-hclust}
luv_clust <- hclust(dist(ggplot2::luv_colours[, 1:3]))
```

```{r 07-class}
class(luv_clust)
```

With the `tidygraph::as_tbl_graph()` function we can transorm the dataset into classes "tbl_graph", "igraph" to make it ready to use for making a visualization of the network data.

```{r 07-luv_graph}
luv_graph <- as_tbl_graph(luv_clust)
luv_graph
class(luv_graph)
```


## Algorithms {-}

The real benefit of networks comes from the different operations that can be performed on them using the underlying structure.

```{r 07-luv_graph2}
luv_graph %>% 
  tidygraph::activate(nodes) %>% 
             mutate(centrality = centrality_pagerank()) %>% 
             arrange(desc(centrality))
```


## Visualizing networks {-}

To visualize the **Network data** we use **{ggraph}**. 

It builds on top of {tidygraph} and {ggplot2} to allow a complete and familiar grammar of graphics for network data. 


## Setting up the visualization {-}

Syntax of **{ggraph}**:

    ggraph() %>%
       ggraph::geom_<functions>

it will choose an appropriate layout based on the type of graph you provide.

[Getting Started guide to layouts](https://ggraph.data-imaginist.com/articles/Layouts.html)

## Specifying a layout {-}

Basic requirements:

The data frame needs to have at least an x and y column and the same number of rows as there are nodes in the input graph. 

As an **example** we take the `data(highschool, package = "ggraph")` and make a **visualization** of the graph:

```{r include = FALSE}
hs_graph <- tidygraph::as_tbl_graph(highschool, directed = FALSE)
```

```{r 07-02, message=FALSE, warning=FALSE, paged.print=FALSE}
library(ggraph)
ggraph(hs_graph) + 
  geom_edge_link() + 
  geom_node_point()
```

## {-}

A second **example** with more features:

```{r 07-03}
hs_graph <- hs_graph %>%
  tidygraph::activate(edges) %>% 
             mutate(edge_weights = runif(n()))

ggraph(hs_graph, layout = "stress", weights = edge_weights) + 
  geom_edge_link(aes(alpha = edge_weights)) + 
  geom_node_point() + 
  scale_edge_alpha_identity()
```


## Many possible layouts {-} 

There are many different possible [layouts](https://www.data-imaginist.com/2017/ggraph-introduction-layouts/).

DRL force-directed graph layout from [igraph](https://igraph.org/r/doc/layout_with_drl.html):

```{r 07-03b}
layout <- ggraph::create_layout(hs_graph, layout = 'drl')

ggraph(layout) + 
  geom_edge_link() + 
  geom_node_point()
```

## {-}

Instead of {tidygraph} we use {igraph}, with layout = "kk": layout.kamada.kawai

```{r 07-04, message=FALSE, warning=FALSE, paged.print=FALSE}
library(ggraph)
library(igraph)

hs_graph2 <- igraph::graph_from_data_frame(highschool)

layout <- create_layout(hs_graph2, layout = "kk")
class(layout)

ggraph(layout) + 
    geom_edge_link(aes(colour = factor(year))) + 
    geom_node_point() 
```

## More on {igraph} {-}

A very simple example to understand how to make a graph network is from this tutorial: [Networks in igraph](https://kateto.net/netscix2016.html)

To understand a bit more about the graph structure we can use these functions:

```{r 07-04b}
g1 <- igraph::graph(edges=c(1,2, 2,3, 3, 1), n=3, directed=F ) 


E(g1); # access to the edges
V(g1); # the vertics
g1[] # access to the matrix
```


## Circular layouts {-}

Layouts can be **linear** and **circular**.

`coord_polar()` changes the coordinate system, affecting the edges


```{r 07-05}
ggraph(luv_graph, layout = 'dendrogram', circular = TRUE) + 
  geom_edge_link() + 
  coord_fixed()
```


```{r 07-06}
ggraph(luv_graph, layout = 'dendrogram') + 
  geom_edge_link() + 
  coord_polar() + 
  scale_y_reverse()
```

## Drawing nodes {-}

[Nodes](https://ggraph.data-imaginist.com/articles/Nodes.html) are similar to points, but we don't (usually) care explicitly about the x and y values.

      geom_node_<functions>
      geom_node_point()
      geom_node_tile() 


```{r 07-luv_graph_tree}
ggraph(luv_graph, layout = "stress") + 
  geom_edge_link() + 
  geom_node_point(aes(colour = factor(members)),
                  show.legend = F)
```  
    
## Color nodes by centrality {-}

More features could be added to calculate node and edge centrality, such as:
    
* centrality_power()   
* centrality_degree()  

```{r 07-07}
ggraph(luv_graph, layout = "stress") + 
  geom_edge_link() + 
  geom_node_point(aes(colour = centrality_power()))
```

## Making tiles {-}

```{r 07-08, message=FALSE, warning=FALSE, paged.print=FALSE}
ggraph(luv_graph, layout = "treemap") + 
  geom_node_tile(aes(fill = depth))
```


## Drawing edges {-}

`geom_edge_link()` draws straight lines (edges) between the connected nodes
(under the hood: splits up the line in a bunch of small fragments.)


- geom_edge_link()
- geom_edge_link2()
- geom_edge_fan()
- geom_edge_parallel()
- geom_edge_elbow()
    - geom_edge_bend() 
    - geom_edge_diagonal()

[Getting Started guide to edges](https://ggraph.data-imaginist.com/articles/Edges.html)

The `after_stat(index)`:

```{r 07-09}
set.seed(123)
ggraph(hs_graph, layout = "stress") + 
  geom_edge_link(aes(alpha = after_stat(index)))
```

## Interpolating edge colors {-}

Let's make a graph artificially with `tidygraph::play_gnp()` and edit it.

Note use of `.N$class[from]` even when edges are `activate`d.

```{r 07-10}
graph <- tidygraph::play_gnp(n = 10, p = 0.2) %>% 
                    activate(nodes) %>% 
                    mutate(class = sample(letters[1:4], 
                                          n(), replace = TRUE)) %>% 
                    activate(edges) %>% 
                    arrange(.N()$class[from])
```

Interpolating colors between nodes: 

```{r}
ggraph(graph, layout = "stress") + 
  geom_edge_link2(
    aes(colour = node.class), 
    width = 3,
    lineend = "round")
```

"Edge geoms have access to the variables of the terminal nodes through specially prefixed variables."

## Other types of edges {-}

```{r 07-11}
ggraph(hs_graph, layout = "stress") + 
  geom_edge_parallel()
```

## Trees and specifically **dendrograms**: {-}

```{r 07-12}
ggraph(luv_graph, layout = "dendrogram", height = height) + 
  geom_edge_elbow()
```

## Clipping edges around nodes {-}

Example: using arrows to show directionality of edges

```{r 07-13}
set.seed(1011)
ggraph(graph, layout = "stress") + 
  geom_edge_link(
    arrow = arrow(), 
    start_cap = circle(5, "mm"),
    end_cap = circle(5, "mm")
  ) + 
  geom_node_point(aes(colour = class), size = 8)
```


## An edge is not always a line {-}

Nodes and edges are abstract concepts and can be visualized in a multitude of ways.

-  geom_edge_point()

Recall: **adjacency matrix**

```{r 07-14}
ggraph(hs_graph, layout = "matrix", sort.by = node_rank_traveller()) + 
  geom_edge_point()
```

## Faceting {-}

* facet_nodes() 
* facet_edges()
* facet_graph()

```{r 07-15}
ggraph(hs_graph, layout = "stress") + 
  geom_edge_link() + 
  geom_node_point() + 
  facet_edges(~year)
```

This is very useful for e.g. multilayer networks!

## Conclusions {-}

* Network data is awkward to represent in tidy format
* `{tidygraph}` uses linked data frames of **nodes** and **edges**
* Special verbs for graph manipulation
* Layouts can be passed as strings or objects
* Edges can have many possible representations
* `{igraph}` can also be used for graph visualization, through a base R plotting framework. 

## Resources {-}

- [tidygraph website](https://tidygraph.data-imaginist.com)
- [Data Imaginist](https://ggraph.data-imaginist.com)
- [Imaginist layouts](https://www.data-imaginist.com/2017/ggraph-introduction-layouts/)
- [Network analysis with r](https://www.jessesadler.com/post/network-analysis-with-r/)
- [R and igraph](https://igraph.org/r/)
- [Getting Started guide to layouts](https://ggraph.data-imaginist.com/articles/Layouts.html)
- [Getting Started guide to nodes](https://ggraph.data-imaginist.com/articles/Nodes.html)
- [Getting Started guide to edges](https://ggraph.data-imaginist.com/articles/Edges.html)



## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/RiACZPhm53Y")`

<details>
  <summary> Meeting chat log </summary>
```
00:09:20	priyanka gagneja:	sorry everyone I just joined
00:09:38	Federica Gazzelloni:	Hello!
00:10:02	priyanka gagneja:	and will probably be a little in and out .. got a not so happy baby today at home
00:18:11	Stan Piotrowski:	I need to take off for a conflict that just came up.  Catch up with you all on slack!
00:20:46	SriRam:	It is the image product ID
00:21:07	SriRam:	All the IDE’s
00:21:50	Kent Johnson:	The IDE codes are defined here: https://ropensci.github.io/bomrang/reference/get_available_imagery.html
00:27:32	SriRam:	The process is called geo-referencing
00:27:58	SriRam:	And image is called a geo-referenced image
00:31:33	SriRam:	Yes, it is a reference system
00:31:38	SriRam:	A coordinate reference
00:33:04	Federica Gazzelloni:	this is the bit that makes the reference: crs = st_crs(sat_vis)
00:49:27	Jiwan Heo:	something just came up, and have to leave. See you all next week!
00:59:58	priyanka gagneja:	I am signing off now , can someone please address and sign off on my behalf. I will send a msg later on slack
```
</details>

`r knitr::include_url("https://www.youtube.com/embed/SV-fN7qYPso")`

<details>
  <summary> Meeting chat log </summary>
```
00:15:36	Ryan S:	https://www.youtube.com/playlist?list=PLkrJrLs7xfbWjD2rp3pIV85lby-tR3Cnu
00:15:48	Lydia Gibson:	Thanks Ryan!
00:16:00	Ryan S:	link to a very good basic tutorial on simple features
00:53:58	Lydia Gibson:	What is GPU?
00:54:23	Ryan S:	GPU is the graphics processing unit (I think)
00:54:31	Lydia Gibson:	Thank you
00:54:32	Ryan S:	it's the part that "draws" on your screen
00:54:43	Lydia Gibson:	Oh okay
00:54:56	Ryan S:	versus the CPU that does calculations
00:55:48	SriRam:	For 2D and non texture plots, I think it is more a RAM issue
00:55:59	Ryan Metcalf:	Oh. I’m so sorry for using Acronyms! Ryan S. is correct. The balance I’m asking Federica is related….”Can I use a slow Laptop or do I have to use a super computer with massive Video card to render these types of graphical objects.
00:56:23	Lydia Gibson:	I always thought CPU was synonymous with computer.
00:57:22	Ryan S:	Ryan, just repurpose the GPUs you currently have that are mining crypto
00:57:38	Ryan Metcalf:	:) Agreed!!!
00:57:57	SriRam:	Lol
00:59:14	SriRam:	If you have a spatial network, do not miss out on “sfnetworks” package
01:00:07	Federica Gazzelloni:	https://kateto.net/netscix2016.html
01:00:14	Federica Gazzelloni:	https://www.data-imaginist.com/2017/ggraph-introduction-layouts/
01:00:24	Federica Gazzelloni:	https://www.hcbravo.org/networks-across-scales/misc/tidygraph.nb.html
01:00:41	Federica Gazzelloni:	https://igraph.org/r/doc/layout_with_drl.html
01:00:48	Federica Gazzelloni:	https://tidygraph.data-imaginist.com/reference/index.html#section-misc
01:00:57	Federica Gazzelloni:	https://ggraph.data-imaginist.com/articles/Layouts.html
01:01:52	Federica Gazzelloni:	https://web.stanford.edu/class/bios221/book/Chap-Graphs.html
https://github.com/jtichon/ModernStatsModernBioJGT/tree/master/data
https://simplemaps.com/data/world-cities
01:07:00	SriRam:	Tidy is I think , Hadley definition, variable is a column, sample point is a row
01:07:36	SriRam:	Sorry my microphone does not work since a few sessions now 🙁
01:07:50	Ryan S:	borders on "marketing" to some degree.  :)
01:08:08	SriRam:	Sfnetwork is not for graphs, it is more for spatial operations
```
</details>
