# Annotations

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE, 
  fig.showtext = TRUE
)

```


```{r load packages, include=FALSE}
library(tidyverse)
library(palmerpenguins)
library(patchwork)
library(ggthemes)
library(scales)
library(ggtext)


nurses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-05/nurses.csv') %>% janitor::clean_names()
theme_set(theme_minimal())

gd <- penguins %>% 
        filter(!is.na(bill_length_mm)) %>% 
                       filter(!is.na(bill_depth_mm)) %>%  
        group_by(species) %>% 
        summarise(bill_length_mm = mean(bill_length_mm),
                  bill_depth_mm  = mean(bill_depth_mm))

base <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm, color = species, shape = species)) +
  geom_point(alpha = .4) +
    geom_point(data = gd, size = 4) +
theme_bw() +
  guides(color = guide_legend("species"),  shape = guide_legend("species")) +
  labs(
    title = "How does bill Size differ by species?",
    subtitle = "Source: Palmer Station Antarctica LTER and K. Gorman, 2020",
    x = "Length",
    y = "Width",
    caption = "ggplot 2 Book Club" 
  ) +
    theme(plot.title = element_text(hjust = .5, face = "bold")) +
   theme(plot.subtitle = element_text(hjust = .5, size = 9)) +
  theme_fivethirtyeight()

astronauts <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-14/astronauts.csv")
  
```


***Learning Objectives*** 

- Plot and Axis Titles; Providing context for the visual, and changing the look of plot elements and overall appearance 

- Text Labels; mapping text from data or having text appear on graphs as data

- Building Custom Annotations; how to write summaries, context, arrows, and textual meta data to graphs 

- Direct Labeling and Faceting; related packages for special issues such as higlighting, textboxes, html text


```{r include=FALSE}

g <- 
nurses %>% 
  group_by(year) %>% 
  filter(state %in% c("Minnesota", "Wisconsin", "Iowa", "North Dakota", "Illinois", "Indiana", "Kansas", "Michigan", "Missouri", "Nebraska", "Ohio")) %>% 
  ggplot(aes(year, annual_salary_median, color = state, )) +
  geom_line() +
  labs(
    title = "Annual Median RN Salary by Midwestern State"
  ) +
   theme(legend.position = "none") +
    geom_vline(xintercept = c(2007, 2009), size = 1.5,
             color = "darkgoldenrod1", linetype = "dashed") +
   gghighlight::gghighlight(state == c("Minnesota", "Wisconsin", "Iowa")) +
  theme_economist() +
  scale_color_economist(name = NULL) +
  theme(axis.title = element_blank()) +
  scale_y_continuous(labels = comma_format())

curved_lines <- base +
  annotate(
    geom = "curve", x = 53, y = 20, xend = 49, yend = 18.5, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
   annotate(geom = "text", x = 53.1, y = 20, label = "Average Chinstrap", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 35, y = 20, xend = 38, yend = 18.5, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 32, y = 20.3, label = "Average Adelie", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 53, y = 15, xend = 48, yend = 15, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 53, y = 15.3, label = "Average Gentoo", hjust = "left", size = 4, color = "darkcyan") 

lab_long <- "**The Great Recession** <br><b style='font-size:10pt;color:steelblue;'> Minnesota's RN Annual Salaries increased during the great receision and then completely flatted out before rising again after 2015"

g_text_box <- g +
  geom_textbox(aes(x = 2015, y = 40000, label = lab_long),
               width = unit(15, "lines"), stat = "unique")

weight <- base + 
  geom_label(aes(label = body_mass_g))


```

## Introduction 

```{r out.width = "80%", fig.asp=1}

(curved_lines | weight) / g 
```
]

---

Packages
- ggtext 
- ggtheme
- gghighlight
- palmerpenguins 
- ggrepel
- grid

Functions
- geom_text
- geom_label
- theme(plot.title = element_text())
- geom = "curve"
- geom_vline 

Resource 
- [A ggplot Tutorial For Beautiful Plotting in R](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/#text) by Cedric Scherer August 5, 2019. 

Annotation Definitions 


- "Conceptually, an annotation supplies metadata for the plot: that is, it provides additional information about the data being displayed. From a practical standpoint, however, metadata is just another form of data. Because of this, the annotation tools in ggplot2 reuse the same geoms that are used to create other plots." 

Wickham, H., Navarro, N., & Lin Pedersen, T. (2016). [Ggplot2: Elegant graphics for data analysis (Second ed.) Springer.](https://ggplot2-book.org/annotations.html)


- "[Annotation] concerns judging the level of assistance an audience may require in order to understand the background, function and purpose of a project, as well as what guidance needs to be provided to help viewers perceive and interpret the data representations."

Kirk, Andy. Data Visualisation (p. 231). SAGE Publications. Kindle Edition. 


## Plot and Axis Titles

```{r echo=TRUE, warning=FALSE, fig.align='center', fig.show='hide'}
base <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm, 
                        color = species, shape = species)) +
  geom_point(alpha = .4) +
  geom_point(data = gd, size = 4) +
theme_bw() +
labs(
title = "How does Bill Size Differ by species?",
subtitle = "Source: Palmer Station Antarctica LTER and K. Gorman, 2020",
x = "*Length*",
y = "Width",
caption = "ggplot 2 Book Club") +
theme(plot.title = element_text(color = "midnightblue", 
      hjust = .5, face = "bold")) +
theme(plot.subtitle = element_text(hjust = .5, size = 9)) + 
theme(axis.title.x = ggtext::element_markdown()) 
    ```


```{r echo=FALSE, warning =FALSE, out.width="70%", fig.align='center'}

gd <- penguins %>% 
        filter(!is.na(bill_length_mm)) %>% 
                       filter(!is.na(bill_depth_mm)) %>%  
        group_by(species) %>% 
        summarise(bill_length_mm = mean(bill_length_mm),
                  bill_depth_mm  = mean(bill_depth_mm))

base <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm, 
              color = species, shape = species)) +
  geom_point(alpha = .4) +
    geom_point(data = gd, size = 4) +
theme_bw() +
  guides(color = guide_legend("species"),  shape = guide_legend("species")) +
  labs(
    title = "How does Bill Size Differ by species?",
    subtitle = "Source: Palmer Station Antarctica LTER and K. Gorman, 2020",
    x = "*Length*",
    y = "Width",
    caption = "ggplot 2 Book Club" 
  ) +
    theme(plot.title = element_text(color = "midnightblue", hjust = .5, face = "bold")) +
   theme(plot.subtitle = element_text(hjust = .5, size = 9)) + 
   theme(axis.title.x = ggtext::element_markdown()) 
base
```

- line breaks \n
- quote() for mathamatical expressions. ?plotmath
- removing labels two ways: labs(x = "") and labs(x = NULL)


## Text labels


8.2 Text labels - geom_text() 
- geom_text() adds label text to the x and y coorindates of a graph such as name instead of a circle in a scatter plot. 

- Change the font with the family aesthetic

- The packages showtext and extrafont can help with handling fonts across differnet devises

- Change the fontface aesthetic for plain, bold, or italic "faces". 

- Alignment: hjust (“left”, “center”, “right”, “inward”, “outward”) and vjust (“bottom”, “middle”, “top”, “inward”, “outward”) aesthetics.

- vjust = "inward", hjust = "inward" ensures labels stay in the plot 

- geom_text(aes(label = text), vjust = "inward", hjust = "inward")
---

```{r echo=TRUE, fig.align="center"}
df <- data.frame(x = 1, y = 3:1, face = c("plain", "bold", "italic"))
ggplot(df, aes(x, y)) + 
  geom_text(aes(label = face, fontface = face, ), 
  vjust = "inward", hjust = "inward", size = 20, angle = 10) 

```

```{r, echo=TRUE, fig.align="center", out.width="70%"}
base + 
  geom_text(aes(label = body_mass_g), check_overlap = TRUE)
```

```{r echo=TRUE, fig.align="center", out.width="70%"}
base + 
  geom_label(aes(label = body_mass_g))
```


```{r echo=TRUE, fig.align="center", out.width="70%"}
ggplot(mpg, aes(displ, hwy)) + 
  geom_text(aes(label = model)) + 
  xlim(1, 8)
```

```{r echo=TRUE, fig.align="center", out.width="70%"}
ggplot(mpg, aes(displ, hwy)) + 
  geom_text(aes(label = model)) + 
  xlim(1, 8)
ggplot(mpg, aes(displ, hwy)) + 
  geom_text(aes(label = model), check_overlap = TRUE) + 
  xlim(1, 8)
```

```{r echo=TRUE, fig.align="center", out.width="70%"}
library(ggrepel)
ggplot(mpg, aes(displ, hwy)) + 
  geom_text_repel(aes(label = model)) + 
  xlim(1, 8)
```



```{r,echo = TRUE, fig.show = 'hide', fig.align="center", out.width="70%"}
label <- data.frame(
  waiting = c(55, 80), 
  eruptions = c(2, 4.3), 
  label = c("peak one", "peak two")
)

ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_tile(aes(fill = density)) + 
  geom_label(data = label, aes(label = label))

```

```{r echo=FALSE, fig.align="center"}
label <- data.frame(
  waiting = c(55, 80), 
  eruptions = c(2, 4.3), 
  label = c("peak one", "peak two")
)

ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_tile(aes(fill = density)) + 
  geom_label(data = label, aes(label = label))

```



geom_label
```{r, fig.align="center"}
base +
  geom_label(aes(x = 42, y = 20, label = "The Adelie species is on all 3 islands"), stat = "unique", size = 5, color = "darkcyan")
```

## Annotations


8.3 Annotations - ggplot2 annotation options 
- geom_text and geom_label

- geom_rect()

- geom_line(), geom_path(), geom_segment(), arrow()

- geom_vline(), geom_hline(), geom_abline()

- annotate() which can be used in combination with arrow()



```{r, fig.align="center", echo=TRUE}
base + annotate(
    geom = "text", x = 42, y = 20, label = "The Adelie species is on all 3 islands", size = 5, color = "darkcyan")
```

Arrows Code 
```{r, echo = TRUE, fig.show = 'hide', fig.align="center"}
base +
  annotate(
    geom = "curve", x = 53, y = 20, xend = 49, yend = 18.5, 
      curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
   annotate(geom = "text", x = 53.1, y = 20, 
    label = "Average Chinstrap", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 35, y = 20, xend = 38, yend = 18.5, 
    curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 32, y = 20.3, 
           label = "Average Adelie", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 53, y = 15, xend = 48, yend = 15, 
    curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 53, y = 15.3, 
           label = "Average Gentoo", hjust = "left", size = 4, color = "darkcyan") 

```

Arrows Plot 

```{r,echo=TRUE, fig.align='center'}
base +
  annotate(
    geom = "curve", x = 53, y = 20, xend = 49, yend = 18.5, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
   annotate(geom = "text", x = 53.1, y = 20, label = "Average Chinstrap", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 35, y = 20, xend = 38, yend = 18.5, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 32, y = 20.3, label = "Average Adelie", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 53, y = 15, xend = 48, yend = 15, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 53, y = 15.3, label = "Average Gentoo", hjust = "left", size = 4, color = "darkcyan") +
    theme(legend.position = "none")
```


```{r echo=TRUE, fig.align="center"}
 astronauts %>% 
  filter(nationality %in% c("U.S.","Australia", "U.K.", "U.S.S.R/Russia", "Japan")) %>% 
  ggplot(aes(x = nationality, y = hours_mission, color = hours_mission)) +
  coord_flip() +
    geom_point(size = 4, alpha = 0.15) +
    geom_boxplot(color = "gray60", outlier.alpha = 0) +
    stat_summary(fun = mean, geom = "point", size = 5, color = "dodgerblue") +
   annotate(
   geom = "curve", x = 3.8, y = 2500, xend = 4, yend = 650,
   curvature = .3, arrow = arrow(length = unit(2, "mm"))
) +
 annotate(
    "text", x = 3.7, y = 2500,
    label = "The U.S. Mean Hours Mission", size = 2.7) +
   annotate(
   geom = "curve", x = 4.7, y = 4200, xend = 5, yend = 2800,
   curvature = .3, arrow = arrow(length = unit(2, "mm"))
) +
 annotate(
    "text", x = 4.5, y = 3700,
    label = "The interquartile range, between 25% and 75% of values", size = 2.8) +
    annotate(
   geom = "curve", x = 1, y = 3800, xend = 1, yend = 900,
   curvature = .3, arrow = arrow(length = unit(2, "mm"))
) +
 annotate(
    "text", x = .8, y = 3000,
    label = "Australian Astronaut Andrew S. W. Thomas 
    completed missions in 1983, 1998, 2001, 2005 and is now retired", size = 2.8) +
  scale_color_viridis_c() +
  scale_y_continuous(limits = c(0, 5000)) +
  labs(title = "Length of Astronaut Missions in hours",
       subtitle = "A Study was conducted on the effects of space on various individuals",
       caption = "Source: TidyTuesday 2020 week 29 \n inspired by plots in The Evolution of a ggplot (ep1) by Cedric Scherer") +
    theme_fivethirtyeight() +
  theme(legend.position = "none") +
  theme(plot.title = element_text(hjust = .5)) +
    theme(plot.subtitle = element_text(hjust = .5)) 
```

## Directlabels Package 

- Place labels closer to the data than legends

- ggforce()

- gghighlight()

```{r}
base +
    directlabels::geom_dl(aes(label = species), method = "smart.grid") +
   theme(legend.position = "none")
```

Base Code Nurse Salary
```{r,echo=TRUE}
library(ggthemes)
library(scales)
library(ggthemes)
library(scales)
g <- 
nurses %>% 
  group_by(year) %>% 
  filter(state %in% c("Minnesota", "Wisconsin", "Iowa", "North Dakota", "Illinois", "Indiana", "Kansas", "Michigan", "Missouri", "Nebraska", "Ohio")) %>% 
  ggplot(aes(year, annual_salary_median, color = state, )) +
  geom_line() +
  labs(
    title = "Annual Median RN Salary by Midwestern State"
  ) +
   theme(legend.position = "none") +
    geom_vline(xintercept = c(2007, 2009), size = 1.5,
             color = "darkgoldenrod1", linetype = "dashed") +
   gghighlight::gghighlight(state == c("Minnesota", "Wisconsin", "Iowa")) +
  theme_economist() +
  scale_color_economist(name = NULL) +
  theme(axis.title = element_blank()) +
  scale_y_continuous(labels = comma_format())
```

```{r, fig.align = 'center', out.width = "70%"}

g 
```

gghighlight and facets

```{r, echo = TRUE, fig.show='hide'}
base + 
  gghighlight::gghighlight() +
  facet_wrap(~ species)
```


```{r echo=FALSE}
base + 
  gghighlight::gghighlight() +
  facet_wrap(~ species)
```

examples in geom_richtext

```{r echo=TRUE, fig.show='hide'}
library(ggtext)

lab_html <- "&#9733; geom_richtext can modify with hmtl"

g +
  geom_richtext(aes(x = 2010, y = 50000, label = lab_html),
                stat = "unique", angle = 30, color = "white", fill = "steelblue")
```

```{r echo = FALSE}
library(ggtext)

lab_html <- "&#9733; geom_richtext can modify with hmtl"

g +
  geom_richtext(aes(x = 2010, y = 50000, label = lab_html),
                stat = "unique", angle = 30, color = "white", fill = "steelblue")
```


geom_textbox
```{r echo=TRUE, fig.align='center', fig.show='hide'}
lab_long <- "**The Great Recession** <br><b style='font-size:10pt;color:steelblue;'> Minnesota's RN Annual Salaries increased during the great receision and then completely flatted out before rising again after 2015"

g +
  geom_textbox(aes(x = 2015, y = 40000, label = lab_long),
               width = unit(15, "lines"), stat = "unique")
```

```{r echo=FALSE, fig.align='center'}
lab_long <- "**The Great Recession** <br><b style='font-size:10pt;color:steelblue;'> Minnesota's RN Annual Salaries increased during the great receision and then completely flatted out before rising again after 2015"

g +
  geom_textbox(aes(x = 2015, y = 40000, label = lab_long),
               width = unit(15, "lines"), stat = "unique")
```

## Faceting Annotations


```{r echo=TRUE, fig.align='center'}
g +
    facet_wrap(~state, scales = "free_x")
```


Grid package scales coordinates between 0 and 1
```{r, echo = TRUE, fig.show='hide', fig.align="center"}
library(grid)
my_grob <- grobTree(textGrob("Great Recession",
                             x = .2, y = .9, hjust = 0,
                             gp = gpar(col = "black",
                                       fontsize = 10,
                                       fontface = "bold")))

g +
    annotation_custom(my_grob) +
    facet_wrap(~state, scales = "free_x") 
```

```{r, echo=FALSE, out.width = "70%", fig.align="center"}
library(grid)
my_grob <- grobTree(textGrob("Great Recession",
                             x = .2, y = .9, hjust = 0,
                             gp = gpar(col = "black",
                                       fontsize = 10,
                                       fontface = "bold")))

g +
    annotation_custom(my_grob) +
    facet_wrap(~state, scales = "free_x") 
```


## Resources
- [ggplot 2 book](https://ggplot2-book.org/annotations.html) chapter 8 annotations 
- [A ggplot Tutorial For Beautiful Plotting in R](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/#text)  by Cedric Scherer August 5, 2019 
- [The Evolution of a ggplot (EP.1)](https://www.cedricscherer.com/2019/05/17/the-evolution-of-a-ggplot-ep.-1/) by Cedric Scherer
- [Introduction to gghighlight](https://cran.r-project.org/web/packages/gghighlight/vignettes/gghighlight.html) by Hiroaki Yutani 2021-06-05


