# Annotations

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  message = FALSE, 
  warning = FALSE, 
  fig.showtext = TRUE
)

```


```{r 08-load packages, include=FALSE}
library(tidyverse)
library(palmerpenguins)
library(patchwork)
library(ggthemes)
library(scales)
library(ggtext)


nurses <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-10-05/nurses.csv') %>% janitor::clean_names()
theme_set(theme_minimal())

gd <- penguins %>% 
        filter(!is.na(bill_length_mm)) %>% 
                       filter(!is.na(bill_depth_mm)) %>%  
        group_by(species) %>% 
        summarise(bill_length_mm = mean(bill_length_mm),
                  bill_depth_mm  = mean(bill_depth_mm))

base <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm, color = species, shape = species)) +
  geom_point(alpha = .4) +
    geom_point(data = gd, size = 4) +
theme_bw() +
  guides(color = guide_legend("species"),  shape = guide_legend("species")) +
  labs(
    title = "How does bill Size differ by species?",
    subtitle = "Source: Palmer Station Antarctica LTER and K. Gorman, 2020",
    x = "Length",
    y = "Width",
    caption = "ggplot 2 Book Club" 
  ) +
    theme(plot.title = element_text(hjust = .5, face = "bold")) +
   theme(plot.subtitle = element_text(hjust = .5, size = 9)) +
  theme_fivethirtyeight()

astronauts <- read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-07-14/astronauts.csv")
  
```


***Learning Objectives*** 

- Plot and Axis Titles; Providing context for the visual, and changing the look of plot elements and overall appearance 

- Text Labels; mapping text from data or having text appear on graphs as data

- Building Custom Annotations; how to write summaries, context, arrows, and textual meta data to graphs 

- Direct Labeling and Faceting; related packages for special issues such as higlighting, textboxes, html text


```{r include=FALSE}

g <- 
nurses %>% 
  group_by(year) %>% 
  filter(state %in% c("Minnesota", "Wisconsin", "Iowa", "North Dakota", "Illinois", "Indiana", "Kansas", "Michigan", "Missouri", "Nebraska", "Ohio")) %>% 
  ggplot(aes(year, annual_salary_median, color = state, )) +
  geom_line() +
  labs(
    title = "Annual Median RN Salary by Midwestern State"
  ) +
   theme(legend.position = "none") +
    geom_vline(xintercept = c(2007, 2009), size = 1.5,
             color = "darkgoldenrod1", linetype = "dashed") +
   gghighlight::gghighlight(state == c("Minnesota", "Wisconsin", "Iowa")) +
  theme_economist() +
  scale_color_economist(name = NULL) +
  theme(axis.title = element_blank()) +
  scale_y_continuous(labels = comma_format())

curved_lines <- base +
  annotate(
    geom = "curve", x = 53, y = 20, xend = 49, yend = 18.5, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
   annotate(geom = "text", x = 53.1, y = 20, label = "Average Chinstrap", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 35, y = 20, xend = 38, yend = 18.5, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 32, y = 20.3, label = "Average Adelie", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 53, y = 15, xend = 48, yend = 15, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 53, y = 15.3, label = "Average Gentoo", hjust = "left", size = 4, color = "darkcyan") 

lab_long <- "**The Great Recession** <br><b style='font-size:10pt;color:steelblue;'> Minnesota's RN Annual Salaries increased during the great receision and then completely flatted out before rising again after 2015"

g_text_box <- g +
  geom_textbox(aes(x = 2015, y = 40000, label = lab_long),
               width = unit(15, "lines"), stat = "unique")

weight <- base + 
  geom_label(aes(label = body_mass_g))


```

## Introduction 

```{r out.width = "80%", fig.asp=1}

(curved_lines | weight) / g 
```
]

---

Packages
- ggtext 
- ggtheme
- gghighlight
- palmerpenguins 
- ggrepel
- grid

Functions
- geom_text
- geom_label
- theme(plot.title = element_text())
- geom = "curve"
- geom_vline 

Resource 
- [A ggplot Tutorial For Beautiful Plotting in R](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/#text) by Cedric Scherer August 5, 2019. 

Annotation Definitions 


- "Conceptually, an annotation supplies metadata for the plot: that is, it provides additional information about the data being displayed. From a practical standpoint, however, metadata is just another form of data. Because of this, the annotation tools in ggplot2 reuse the same geoms that are used to create other plots." 

Wickham, H., Navarro, N., & Lin Pedersen, T. (2016). [Ggplot2: Elegant graphics for data analysis (Second ed.) Springer.](https://ggplot2-book.org/annotations.html)


- "[Annotation] concerns judging the level of assistance an audience may require in order to understand the background, function and purpose of a project, as well as what guidance needs to be provided to help viewers perceive and interpret the data representations."

Kirk, Andy. Data Visualisation (p. 231). SAGE Publications. Kindle Edition. 


## Plot and Axis Titles

```{r echo=TRUE, warning=FALSE, fig.align='center', fig.show='hide'}
base <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm, 
                        color = species, shape = species)) +
  geom_point(alpha = .4) +
  geom_point(data = gd, size = 4) +
theme_bw() +
labs(
title = "How does Bill Size Differ by species?",
subtitle = "Source: Palmer Station Antarctica LTER and K. Gorman, 2020",
x = "*Length*",
y = "Width",
caption = "ggplot 2 Book Club") +
theme(plot.title = element_text(color = "midnightblue", 
      hjust = .5, face = "bold")) +
theme(plot.subtitle = element_text(hjust = .5, size = 9)) + 
theme(axis.title.x = ggtext::element_markdown()) 
```


```{r echo=FALSE, warning =FALSE, out.width="70%", fig.align='center'}

gd <- penguins %>% 
        filter(!is.na(bill_length_mm)) %>% 
                       filter(!is.na(bill_depth_mm)) %>%  
        group_by(species) %>% 
        summarise(bill_length_mm = mean(bill_length_mm),
                  bill_depth_mm  = mean(bill_depth_mm))

base <- ggplot(penguins, aes(bill_length_mm, bill_depth_mm, 
              color = species, shape = species)) +
  geom_point(alpha = .4) +
    geom_point(data = gd, size = 4) +
theme_bw() +
  guides(color = guide_legend("species"),  shape = guide_legend("species")) +
  labs(
    title = "How does Bill Size Differ by species?",
    subtitle = "Source: Palmer Station Antarctica LTER and K. Gorman, 2020",
    x = "*Length*",
    y = "Width",
    caption = "ggplot 2 Book Club" 
  ) +
    theme(plot.title = element_text(color = "midnightblue", hjust = .5, face = "bold")) +
   theme(plot.subtitle = element_text(hjust = .5, size = 9)) + 
   theme(axis.title.x = ggtext::element_markdown()) 
base
```

- line breaks \n
- quote() for mathamatical expressions. ?plotmath
- removing labels two ways: labs(x = "") and labs(x = NULL)


## Text labels


8.2 Text labels - geom_text() 
- geom_text() adds label text to the x and y coorindates of a graph such as name instead of a circle in a scatter plot. 

- Change the font with the family aesthetic

- The packages showtext and extrafont can help with handling fonts across differnet devises

- Change the fontface aesthetic for plain, bold, or italic "faces". 

- Alignment: hjust (“left”, “center”, “right”, “inward”, “outward”) and vjust (“bottom”, “middle”, “top”, “inward”, “outward”) aesthetics.

- vjust = "inward", hjust = "inward" ensures labels stay in the plot 

- geom_text(aes(label = text), vjust = "inward", hjust = "inward")
---

```{r echo=TRUE, fig.align="center"}
df <- data.frame(x = 1, y = 3:1, face = c("plain", "bold", "italic"))
ggplot(df, aes(x, y)) + 
  geom_text(aes(label = face, fontface = face, ), 
  vjust = "inward", hjust = "inward", size = 20, angle = 10) 

```

```{r, echo=TRUE, fig.align="center", out.width="70%"}
base + 
  geom_text(aes(label = body_mass_g), check_overlap = TRUE)
```

```{r echo=TRUE, fig.align="center", out.width="70%"}
base + 
  geom_label(aes(label = body_mass_g))
```


```{r echo=TRUE, fig.align="center", out.width="70%"}
ggplot(mpg, aes(displ, hwy)) + 
  geom_text(aes(label = model)) + 
  xlim(1, 8)
```

```{r echo=TRUE, fig.align="center", out.width="70%"}
ggplot(mpg, aes(displ, hwy)) + 
  geom_text(aes(label = model)) + 
  xlim(1, 8)
ggplot(mpg, aes(displ, hwy)) + 
  geom_text(aes(label = model), check_overlap = TRUE) + 
  xlim(1, 8)
```

```{r echo=TRUE, fig.align="center", out.width="70%"}
library(ggrepel)
ggplot(mpg, aes(displ, hwy)) + 
  geom_text_repel(aes(label = model)) + 
  xlim(1, 8)
```



```{r,echo = TRUE, fig.show = 'hide', fig.align="center", out.width="70%"}
label <- data.frame(
  waiting = c(55, 80), 
  eruptions = c(2, 4.3), 
  label = c("peak one", "peak two")
)

ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_tile(aes(fill = density)) + 
  geom_label(data = label, aes(label = label))

```

```{r echo=FALSE, fig.align="center"}
label <- data.frame(
  waiting = c(55, 80), 
  eruptions = c(2, 4.3), 
  label = c("peak one", "peak two")
)

ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_tile(aes(fill = density)) + 
  geom_label(data = label, aes(label = label))

```



geom_label
```{r, fig.align="center"}
base +
  geom_label(aes(x = 42, y = 20, label = "The Adelie species is on all 3 islands"), stat = "unique", size = 5, color = "darkcyan")
```

## Annotations


8.3 Annotations - ggplot2 annotation options 
- geom_text and geom_label

- geom_rect()

- geom_line(), geom_path(), geom_segment(), arrow()

- geom_vline(), geom_hline(), geom_abline()

- annotate() which can be used in combination with arrow()



```{r, fig.align="center", echo=TRUE}
base + annotate(
    geom = "text", x = 42, y = 20, label = "The Adelie species is on all 3 islands", size = 5, color = "darkcyan")
```

Arrows Code 
```{r, echo = TRUE, fig.show = 'hide', fig.align="center"}
base +
  annotate(
    geom = "curve", x = 53, y = 20, xend = 49, yend = 18.5, 
      curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
   annotate(geom = "text", x = 53.1, y = 20, 
    label = "Average Chinstrap", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 35, y = 20, xend = 38, yend = 18.5, 
    curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 32, y = 20.3, 
           label = "Average Adelie", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 53, y = 15, xend = 48, yend = 15, 
    curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 53, y = 15.3, 
           label = "Average Gentoo", hjust = "left", size = 4, color = "darkcyan") 

```

Arrows Plot 

```{r,echo=TRUE, fig.align='center'}
base +
  annotate(
    geom = "curve", x = 53, y = 20, xend = 49, yend = 18.5, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
   annotate(geom = "text", x = 53.1, y = 20, label = "Average Chinstrap", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 35, y = 20, xend = 38, yend = 18.5, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 32, y = 20.3, label = "Average Adelie", hjust = "left", size = 4, color = "darkcyan") +
  annotate(
   geom = "curve", x = 53, y = 15, xend = 48, yend = 15, curvature = .3, size = 1, arrow = arrow(length = unit(3, "mm"))
  ) +
  annotate(geom = "text", x = 53, y = 15.3, label = "Average Gentoo", hjust = "left", size = 4, color = "darkcyan") +
    theme(legend.position = "none")
```


```{r echo=TRUE, fig.align="center"}
 astronauts %>% 
  filter(nationality %in% c("U.S.","Australia", "U.K.", "U.S.S.R/Russia", "Japan")) %>% 
  ggplot(aes(x = nationality, y = hours_mission, color = hours_mission)) +
  coord_flip() +
    geom_point(size = 4, alpha = 0.15) +
    geom_boxplot(color = "gray60", outlier.alpha = 0) +
    stat_summary(fun = mean, geom = "point", size = 5, color = "dodgerblue") +
   annotate(
   geom = "curve", x = 3.8, y = 2500, xend = 4, yend = 650,
   curvature = .3, arrow = arrow(length = unit(2, "mm"))
) +
 annotate(
    "text", x = 3.7, y = 2500,
    label = "The U.S. Mean Hours Mission", size = 2.7) +
   annotate(
   geom = "curve", x = 4.7, y = 4200, xend = 5, yend = 2800,
   curvature = .3, arrow = arrow(length = unit(2, "mm"))
) +
 annotate(
    "text", x = 4.5, y = 3700,
    label = "The interquartile range, between 25% and 75% of values", size = 2.8) +
    annotate(
   geom = "curve", x = 1, y = 3800, xend = 1, yend = 900,
   curvature = .3, arrow = arrow(length = unit(2, "mm"))
) +
 annotate(
    "text", x = .8, y = 3000,
    label = "Australian Astronaut Andrew S. W. Thomas 
    completed missions in 1983, 1998, 2001, 2005 and is now retired", size = 2.8) +
  scale_color_viridis_c() +
  scale_y_continuous(limits = c(0, 5000)) +
  labs(title = "Length of Astronaut Missions in hours",
       subtitle = "A Study was conducted on the effects of space on various individuals",
       caption = "Source: TidyTuesday 2020 week 29 \n inspired by plots in The Evolution of a ggplot (ep1) by Cedric Scherer") +
    theme_fivethirtyeight() +
  theme(legend.position = "none") +
  theme(plot.title = element_text(hjust = .5)) +
    theme(plot.subtitle = element_text(hjust = .5)) 
```

## Directlabels Package 

- Place labels closer to the data than legends

- ggforce()

- gghighlight()

```{r}
base +
    directlabels::geom_dl(aes(label = species), method = "smart.grid") +
   theme(legend.position = "none")
```

Base Code Nurse Salary
```{r,echo=TRUE}
library(ggthemes)
library(scales)
library(ggthemes)
library(scales)
g <- 
nurses %>% 
  group_by(year) %>% 
  filter(state %in% c("Minnesota", "Wisconsin", "Iowa", "North Dakota", "Illinois", "Indiana", "Kansas", "Michigan", "Missouri", "Nebraska", "Ohio")) %>% 
  ggplot(aes(year, annual_salary_median, color = state, )) +
  geom_line() +
  labs(
    title = "Annual Median RN Salary by Midwestern State"
  ) +
   theme(legend.position = "none") +
    geom_vline(xintercept = c(2007, 2009), size = 1.5,
             color = "darkgoldenrod1", linetype = "dashed") +
   gghighlight::gghighlight(state == c("Minnesota", "Wisconsin", "Iowa")) +
  theme_economist() +
  scale_color_economist(name = NULL) +
  theme(axis.title = element_blank()) +
  scale_y_continuous(labels = comma_format())
```

```{r, fig.align = 'center', out.width = "70%"}

g 
```

gghighlight and facets

```{r, echo = TRUE, fig.show='hide'}
base + 
  gghighlight::gghighlight() +
  facet_wrap(~ species)
```


```{r echo=FALSE}
base + 
  gghighlight::gghighlight() +
  facet_wrap(~ species)
```

examples in geom_richtext

```{r echo=TRUE, fig.show='hide'}
library(ggtext)

lab_html <- "&#9733; geom_richtext can modify with hmtl"

g +
  geom_richtext(aes(x = 2010, y = 50000, label = lab_html),
                stat = "unique", angle = 30, color = "white", fill = "steelblue")
```

```{r echo = FALSE}
library(ggtext)

lab_html <- "&#9733; geom_richtext can modify with hmtl"

g +
  geom_richtext(aes(x = 2010, y = 50000, label = lab_html),
                stat = "unique", angle = 30, color = "white", fill = "steelblue")
```


geom_textbox
```{r echo=TRUE, fig.align='center', fig.show='hide'}
lab_long <- "**The Great Recession** <br><b style='font-size:10pt;color:steelblue;'> Minnesota's RN Annual Salaries increased during the great receision and then completely flatted out before rising again after 2015"

g +
  geom_textbox(aes(x = 2015, y = 40000, label = lab_long),
               width = unit(15, "lines"), stat = "unique")
```

```{r echo=FALSE, fig.align='center'}
lab_long <- "**The Great Recession** <br><b style='font-size:10pt;color:steelblue;'> Minnesota's RN Annual Salaries increased during the great receision and then completely flatted out before rising again after 2015"

g +
  geom_textbox(aes(x = 2015, y = 40000, label = lab_long),
               width = unit(15, "lines"), stat = "unique")
```

## Faceting Annotations


```{r echo=TRUE, fig.align='center'}
g +
    facet_wrap(~state, scales = "free_x")
```


Grid package scales coordinates between 0 and 1
```{r, echo = TRUE, fig.show='hide', fig.align="center"}
library(grid)
my_grob <- grobTree(textGrob("Great Recession",
                             x = .2, y = .9, hjust = 0,
                             gp = gpar(col = "black",
                                       fontsize = 10,
                                       fontface = "bold")))

g +
    annotation_custom(my_grob) +
    facet_wrap(~state, scales = "free_x") 
```

```{r, echo=FALSE, out.width = "70%", fig.align="center"}
library(grid)
my_grob <- grobTree(textGrob("Great Recession",
                             x = .2, y = .9, hjust = 0,
                             gp = gpar(col = "black",
                                       fontsize = 10,
                                       fontface = "bold")))

g +
    annotation_custom(my_grob) +
    facet_wrap(~state, scales = "free_x") 
```


## Resources
- [ggplot 2 book](https://ggplot2-book.org/annotations.html) chapter 8 annotations 
- [A ggplot Tutorial For Beautiful Plotting in R](https://www.cedricscherer.com/2019/08/05/a-ggplot2-tutorial-for-beautiful-plotting-in-r/#text)  by Cedric Scherer August 5, 2019 
- [The Evolution of a ggplot (EP.1)](https://www.cedricscherer.com/2019/05/17/the-evolution-of-a-ggplot-ep.-1/) by Cedric Scherer
- [Introduction to gghighlight](https://cran.r-project.org/web/packages/gghighlight/vignettes/gghighlight.html) by Hiroaki Yutani 2021-06-05




## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/iKvD5Mas8Ow")`

<details>
  <summary> Meeting chat log </summary>
```
00:10:14	Ed:	Hi everyone. My connection is shaky so if I drop off don’t take it personally. 😇
00:10:32	Michael Haugen:	Thanks for joining us!
00:10:42	Ryan Metcalf:	Great to see you. No worries at all.
00:24:25	Ryan Metcalf:	To support Michael’s quote, I mentioned a Swedish Statician…Hans Rosling. The Gapminder project was his brain child. Great Ted Talks were delivered by the user: https://www.ted.com/speakers/hans_rosling
00:32:42	June Choe:	re: text/font rendering - {ragg} + {systemfonts} is now recommended over {showtext}/{extrafont}!
00:32:59	June Choe:	https://yjunechoe.github.io/posts/2021-06-24-setting-up-and-debugging-custom-fonts/
00:33:39	Federica Gazzelloni:	@June thanks
00:34:28	June Choe:	here's some quotes from Thomas Lin Pedersen (ggplot2 dev) on showtext/extrafont - https://twitter.com/thomasp85/status/1355083725156077571
https://twitter.com/thomasp85/status/1261539815960518656
00:39:31	Ed:	So is it necessary to hard code the locations for those arrows? It won't stop them where it makes sense to go?
00:39:46	Ed:	What about different resolution screens, etc.
00:41:36	Kent Johnson:	Yes, you have to hard-code the arrow start and end.
00:42:09	Ed:	👍
00:42:46	Kent Johnson:	My experience is, it's pretty fiddly to get something really nice. I don't know how plot size / screen resolution affect the arrows.
00:43:42	Ryan Metcalf:	https://fivethirtyeight.com/
00:46:42	June Choe:	linewidth and arrow size would be subject to resolution but not the stard/end points
00:47:07	June Choe:	start/end points are converted to native coordinate units but size is absolute
00:47:46	Ed:	👍
00:48:03	June Choe:	(which is why you should never rely just on plot panel output and always use something like ggsave!)
00:48:58	Ed:	Awesome tip. Could see myself getting frustrated but good to know going into it.
00:49:35	June Choe:	since like an update or two ago, ggsave() started returning the path to the saved image invisibly, so if you
00:50:07	June Choe:	if you're on windows, you can do something like  `system2("open", ggsave("img.png"))` and itll open up the plot after saving it
00:50:27	June Choe:	(open it back up using your system's default photo viewing app)
00:58:21	Ryan Metcalf:	Sheesh! This took me forever to find! I mentioned Arrows outside of a graphic. I was using it with D3 objects (similar to ggplot2). https://github.com/krispo/yarrow
01:01:04	June Choe:	big fan - and you should check out {sinab} as well for a more powerful version of ggtext by the same dev (though this one's heavily experimental and requires Rust) - https://clauswilke.com/sinab/
01:01:18	Michael Haugen:	thanks
01:03:14	June Choe:	the 0-1 coord scale in grid here is called "npc" (Normalized Parent Coordinates)
01:04:21	Ryan Metcalf:	June, you are a wealth of knowledge! 🙂I may ping you outside of Zoom (Slack) for further discussions on Graphical Objects.
01:05:00	Ryan S:	Awesome job Michael!
01:05:12	June Choe:	For sure @Ryan ! Always happy to talk about data viz
01:05:15	June Choe:	and thanks for presenting Michael!
01:05:50	June Choe:	xaringanExtra i think
01:06:22	June Choe:	https://pkg.garrickadenbuie.com/xaringanExtra/#/extra-styles
01:07:31	Federica Gazzelloni:	Thanks Michael
```
</details>
