# Arranging Plots


**Learning Objectives**

- Produce several subplots part of the same main visualization
- A range of packages for providing different approaches to arranging separate plots



## Introduction

This chapter focuses on making more than one plot in one visualization, using the following packages:

1. patchwork
2. cowplot
3. gridExtra
4. ggpubr


## Arranging plots side by side with no overlap

```{r library, message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
```


```{r dataplots, include=FALSE}
p1 <- ggplot(mpg) + 
  geom_point(aes(x = displ, y = hwy))

p2 <- ggplot(mpg) + 
  geom_bar(aes(x = as.character(year), fill = drv), position = "dodge") + 
  labs(x = "year")

p3 <- ggplot(mpg) + 
  geom_density(aes(x = hwy, fill = drv), colour = NA) + 
  facet_grid(rows = vars(drv))

p4 <- ggplot(mpg) + 
  stat_summary(aes(x = drv, y = hwy, fill = drv), geom = "col", fun.data = mean_se) +
  stat_summary(aes(x = drv, y = hwy), geom = "errorbar", fun.data = mean_se, width = 0.5)
```


```{r plot1}
library(patchwork)

p1 + p2 # same as p3 | p4
```

```{r plot2}
p1 + p2 + p3 + p4
```


### Taking control of the layout

```{r plot3}
p1 + p2 + p3 + plot_layout(ncol = 2)
```

```{r plot4}
p1 / p2
```

More compositions:
```{r plot5}
p3 | (p2 / (p1 | p4))
```

### More about layouts

```{r plot6}
layout <- "
AAB
C#B
CDD
"

p1 + p2 + p3 + p4 + plot_layout(design = layout)
```

```{r plot7}
p1 + p2 + p3 + guide_area() + plot_layout(ncol = 2, guides = "collect")
```


This way is possible a custom modification of the theme for one plot or for both.
```{r plot8}
p12 <- p1 + p2
p12[[1]] <- p12[[1]] + theme_light()
p12
```

```{r plot9}
p1 + p4 & 
  scale_y_continuous(limits = c(0, 45)) & 
  theme_minimal()
```

### Plot annotations
```{r plot10}
p34 <- p3 + p4 + 
  plot_annotation(
  title = "A closer look at the effect of drive train in cars",
  caption = "Source: mpg dataset in ggplot2",
  tag_levels = "I") &
  theme_gray(base_family = "mono")

p34
```

```{r plot11}
p34[[2]] <- p34[[2]] + plot_layout(tag_level = "new")
p34 + plot_annotation(tag_levels = c("I", "a"))
```


## Arranging plots on top of each other

It is possible to arrange plots in a way that they are nested to each other, as well as setting the position inside the main plot.

General options are left, right, top, and bottom locations, but more specific locations can be set, such as using: `grid::unit()` (default uses npc units which goes from 0 to 1)

In addition, the location is by default set to the **panel area**, but can be align_to` **plot area**. 

An inset can be placed exactly 15 mm from the top right corner.

```{r plot12}
p1 + 
  inset_element(
    p2, 
    left = 0.4, 
    bottom = 0.4, 
    right = unit(1, "npc") - unit(15, "mm"), 
    top = unit(1, "npc") - unit(15, "mm"),
    align_to = "full"
  )
```

```{r plot13}
p24 <- p2 / p4 + plot_layout(guides = "collect")

p1 + 
  inset_element(p24, left = 0.5, bottom = 0.05, right = 0.95, top = 0.9)  +
  plot_annotation(tag_levels = 'A') & 
  theme_bw()
```



## Extra

**grid and gridExtra packages**
```{r plot14, message=FALSE, warning=FALSE, paged.print=FALSE}
library(gridExtra)
library(grid)

grid.arrange(p1, p2, p3, p4, ncol=2)
```


**cowplot package**
```{r plot15, message=FALSE, warning=FALSE, paged.print=FALSE}
library(cowplot)

plot_grid(p1, p2, 
          labels = c('A', 'B'),
          align = "h",
          label_size = 12,
          label_fontfamily = "serif",
          label_fontface = "plain",
          label_colour = "blue",
          rel_widths = c(1, 2))
```


```{r plot16, message=FALSE, warning=FALSE, paged.print=FALSE}
ggdraw()+
  draw_plot(p3)+
  draw_image("images/p1.png",x=0.4,y=0.4,scale=0.5)+
  draw_label("This is some text",x=0.5,y=0.9)
```



To add a common title we use `ggdraw()
```{r plot17}
plot_row <- plot_grid(p1, p2)

# now add the title
title <- ggdraw() + 
  draw_label("Common title", 
             fontface = 'bold', x = 0, hjust = 0) +
  theme(plot.margin = margin(0, 0, 0, 7))

plot_grid(title, plot_row,
          ncol = 1,
          rel_heights = c(0.1, 1))
```

```{r plot18}
ggdraw()+
  draw_plot(p1,x=0.1,y=0,scale=0.5)+
  draw_plot(p2,x=-0.2,y=0,scale=0.5)+
  draw_label("This plot is made with cowplot",
             x=0.5,y=0.95,
             color="pink",
             fontface="bold",size=20)
```

**ggpubr package**
```{r plot19, message=FALSE, warning=FALSE, paged.print=FALSE}
library(ggpubr)
ggarrange(p1, p2, p3, ncol = 2, nrow = 2,common.legend = TRUE)
```


## Conclusions


[Patchwork - imaginist](https://patchwork.data-imaginist.com/) is one of the packages mentioned in the book, also some other packages provide same results with different approaches. 

### Extra resources:

- [grid and gridExtra](https://cran.r-project.org/web/packages/gridExtra/vignettes/arrangeGrob.html)
- [cowplot](https://wilkelab.org/cowplot/articles/plot_grid.html)
- [ggpubr](https://rpkgs.datanovia.com/ggpubr/reference/ggarrange.html)
