# (PART\*) Scales {-}

# Position scales and axes

**Learning Objectives**

- What are the defining components of a scale?
- When/why does the data need to be transformed for a visualization?
- What are the defining components of an axis?
- What is the relationship between scale and axis?

```{r 10-01}
library(ggplot2)
library(dplyr)
library(stringr) # for demo of labels and some other stuff
```


## Introduction / preliminaries / asides

This chapter introduces position scales and axes. It may also be helpful to understand position scales and axes as position scales and _guides_, because axes they share the same API as guides for non-positional scales like color legends. The parallel will be clearer in the next chapter.

It's worthwhile to read documentations of the `{scales}` package to learn more about scales, since that handles a lot of the (re-)scaling and transformation under the hood. It may be good to start with the [rstudio::conf2020 talk on scales](https://www.rstudio.com/resources/rstudioconf-2020/the-little-package-that-could-taking-visualizations-to-the-next-level-with-the-scales-package/).

```{r 10-02}
knitr::include_url("https://www.rstudio.com/resources/rstudioconf-2020/the-little-package-that-could-taking-visualizations-to-the-next-level-with-the-scales-package/")
```

It should also be noted that there's some discussion about revamping the `scales_*` API. See [issue #4269](https://github.com/tidyverse/ggplot2/issues/4269) and [PR #4271](https://github.com/tidyverse/ggplot2/pull/4271)

Lastly, a small aside on the book's `after_stat()` example  it he intro, continuing nicely from our discussion on ggplot internals last week.

```{r 10-03}
# Grab the layer object created by `geom_histogram()`
histogram_layer <- geom_histogram()
# Check what stat ggproto it uses
class(histogram_layer$stat)[1]
# Confirm that the stat has `after_stat(count)` as the default aes
StatBin$default_aes # or, `histogram_layer$stat$default_aes`
# The stat takes an x or y aesthetic, so it does implicitly maps
# `after_stat(count)` to the unspecified aes
StatBin$required_aes
geom_histogram(aes(x = displ))$mapping
# 'orientation' argument allows horizontal bars without `coord_flip()`
# as of v.3.3.0 (Dec 2020)
ggplot(mpg, aes(y = displ)) +
  geom_histogram(orientation = "y")
```

## 10.1 Numeric

### 10.1.1 Limits

The book doesn't have content for this section (??)

But we know that you can set limits with `xlim()`/`ylim()` or `scale_x|y_*(limits = )`

```{r 10-04}
lim_plot <- ggplot(mtcars, aes(x = hp, y = disp)) +
  geom_point()

lim_plot

lim_plot +
  xlim(0, 500)

# NA to use range of data
lim_plot +
  xlim(0, NA)

# Same thing
lim_plot +
  scale_x_continuous(limits = c(0, NA))
```


### 10.1.2 Out of bounds values

NOTE: A big theme of the `{scales}` package as of v1.1.1 (May 2020) is that they have very transparent function names. For example, the family of functions for Out Of Bounds (oob) handling are all named `oob_*()`. This is an intentional (re-)design of the package to work nicely with autocomplete.

```{r 10-05}
str_subset(ls(envir = asNamespace("scales")), "^oob_")
```

By default, data outside scales are set to `NA`. This is because the `oob` argument is set to `oob_censor()`/`censor()`. Note that oob only applies to continuous scales, since values of a discrete scale form a fixed set.

```{r 10-06}
body(scale_x_continuous)
formals(continuous_scale)$oob
```

Book's examples:

```{r 10-07}
base <- ggplot(mpg, aes(drv, hwy)) + 
  geom_hline(yintercept = 28, colour = "red") + 
  geom_boxplot(alpha = .2) # I set alpha here for a later demo

base
base + coord_cartesian(ylim = c(10, 35))
base + ylim(10, 35)
```

Equivalent solutions with `oob_*()`

```{r 10-08}
# zoom only (keeps out of bounds values for the stat computation)
base + scale_y_continuous(limits = c(10, 35), oob = scales::oob_keep)
# default that removes out of bounds values
base + scale_y_continuous(limits = c(10, 35), oob = scales::oob_censor)
# squish option (plots outliers at the uper limit of y = 35)
base + scale_y_continuous(limits = c(10, 35), oob = scales::oob_squish)
```

You can use oob functions for non-positional scales 

```{r 10-09}
df <- data.frame(x = 1:6, y = 8:13)
base <- ggplot(df, aes(x, y)) + 
  geom_col(aes(fill = x)) +                    # bar chart
  geom_vline(xintercept = 3.5, colour = "red") # for visual clarity only

base
base + scale_fill_gradient(limits = c(1, 3)) # oob = scales::oob_censor
base + scale_fill_gradient(limits = c(1, 3), oob = scales::squish) # scales::oob_squish
```


### 10.1.3 Visual range expansion

Book examples:

```{r 10-10}
f_plot <- ggplot(faithfuld, aes(waiting, eruptions)) + 
  geom_raster(aes(fill = density)) + 
  theme(legend.position = "none")
f_plot

f_plot + 
  scale_x_continuous(expand = c(0,0)) +  # expand = 0
  scale_y_continuous(expand = c(0,0))  # expand = 0
```

With `expansion()` from v3.3.0 (Dec 2020)

```{r 10-11}
formals(expansion)
```

```{r 10-12}
f_plot +
  scale_y_continuous(expand = expansion(mult = 0)) # mult = c(0, 0)
f_plot +
  scale_y_continuous(expand = expansion(mult = 1))
f_plot +
  scale_y_continuous(expand = expansion(mult = c(0, 1)))
f_plot +
  scale_y_continuous(expand = expansion(mult = c(0, 1))) +
  scale_x_continuous(expand = expansion(add = c(0, 10)))
```


### 10.1.4 Exercises


### 10.1.5 Breaks

```{r 10-13}
str_subset(ls(envir = asNamespace("scales")), "^breaks_")
```

Book example:

```{r 10-14}
toy <- data.frame(
  const = 1, 
  up = 1:4,
  txt = letters[1:4], 
  big = (1:4)*1000,
  log = c(2, 5, 10, 2000)
)
toy
#>   const up txt  big  log
#> 1     1  1   a 1000    2
#> 2     1  2   b 2000    5
#> 3     1  3   c 3000   10
#> 4     1  4   d 4000 2000
axs <- ggplot(toy, aes(big, const)) + 
  geom_point() + 
  labs(x = NULL, y = NULL)
axs
axs + scale_x_continuous(breaks = scales::breaks_extended())
axs + scale_x_continuous(breaks = scales::breaks_extended(n = 2))
axs + scale_x_continuous(breaks = NULL)
```

Demo from `{scales}`:

```{r 10-15}
scales::demo_continuous(c(1000, 4000), breaks = scales::breaks_extended())
scales::demo_continuous(c(1000, 4000), breaks = scales::breaks_extended(n = 2))
scales::demo_continuous(c(1000, 4000), NULL)
```

At the vector level:

```{r 10-16}
scales::breaks_extended()(c(1000, 4000))
scales::breaks_extended(n = 2)(c(1000, 4000))
```

Other breaks:

```{r 10-17}
my_range <- c(1, 101)
scales::breaks_extended()(my_range)
scales::breaks_width(width = 10)(my_range)
scales::breaks_pretty(width = 10)(my_range) # pretty(1:101)
scales::breaks_log()(my_range)
```

Debugging arguments in `scale_*()` that take function factories

```{r 10-18, eval = FALSE}
browserer <- function(...) {
  params <- list(...)
  browser()
  if (exists("result")) {
    return(result)
  }
}
axs + scale_x_continuous(breaks = browserer)
```


### 10.1.6 Minor breaks

Book example:

```{r 10-19}
mb <- unique(as.numeric(1:10 %o% 10 ^ (0:3)))
mb

log_base <- ggplot(toy, aes(log, const)) + geom_point()

log_base + scale_x_log10()
log_base + scale_x_log10(minor_breaks = mb)
```

There are also minor break functions:

```{r 10-20}
str_subset(ls(envir = asNamespace("scales")), "^minor_breaks_")
```


### 10.1.7 Labels

```{r 10-21}
str_subset(ls(envir = asNamespace("scales")), "^label_")
```

Book examples:

```{r 10-22}
axs + scale_y_continuous(labels = scales::label_percent())
axs + scale_y_continuous(labels = scales::label_dollar(prefix = "", suffix = "€"))
```

```{r 10-23}
tibble(
  x = c("cat1", "cat2 with a really really realy long name", "cat3"),
  y = 1:3
) %>% 
  ggplot(aes(x, y)) +
  geom_col()
```

```{r 10-24}
tibble(
  x = c("cat1", "cat2 with a really really realy long name", "cat3"),
  y = 1:3
) %>% 
  ggplot(aes(x, y)) +
  geom_col() +
  scale_x_discrete(labels = scales::label_wrap(width = 30))
```


### 10.1.8 Exercises


### 10.1.9 Transformations

Book example:

```{r 10-25}
ggplot(diamonds, aes(price, carat)) + 
  geom_bin2d()
# log transform x and y axes
ggplot(diamonds, aes(price, carat)) + 
  geom_bin2d() + 
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10")
```


> The transformation is carried out by a “transformer”, which describes the transformation, its inverse, and how to draw the labels. You can construct your own transformer using `scales::trans_new()`

Case study: make reversed log x-axis

```{r 10-26}
ggplot(starwars, aes(x = mass)) +
  geom_histogram()

ggplot(starwars, aes(x = mass)) +
  geom_histogram() +
  scale_x_log10()

ggplot(starwars, aes(x = mass)) +
  geom_histogram() +
  scale_x_reverse()
```

```{r 10-27}
scale_x_log10()$trans    # scales::log10_trans()
scale_x_reverse()$trans  # scales::reverse_trans()
```

```{r 10-28}
formals(scales::trans_new)
log10_reverse <- scales::trans_new(
  name = "log-10-reverse",
  transform = function(x) -log(x, 10),
  inverse = function(x) 10^(-x),
  breaks = scales::log10_trans()$breaks,
  minor_breaks = scales::log10_trans()$minor_breaks,
  domain = scales::log10_trans()$domain
)
ggplot(starwars, aes(x = mass)) +
  geom_histogram() +
  scale_x_continuous(trans = log10_reverse)
```

> Regardless of which method you use, the transformation occurs before any statistical summaries. To transform after statistical computation use `coord_trans()`

From the docs:

```{r 10-29}
ggplot(diamonds, aes(carat, price)) +
  geom_point() +
  scale_x_log10() +
  scale_y_log10()

ggplot(diamonds, aes(carat, price)) +
  geom_point() +
  coord_trans(x = "log10", y = "log10")
```

Example where stat transformation matters:

```{r 10-30}
trans_plot <- ggplot(mpg, aes(drv, hwy)) +
  geom_boxplot()
trans_plot +
  scale_y_log10()
trans_plot +
  coord_trans(y = "log10") # scales::log10_trans()
```


```{r 10-31}
layer_data(trans_plot) %>% 
  select(x, starts_with("y"))
layer_data(trans_plot + scale_y_log10()) %>% 
  select(x, starts_with("y"))
layer_data(trans_plot + coord_trans(y = "log10")) %>% 
  select(x, starts_with("y"))
```


### ASIDE - A little more on transformations

`transform()` method of the [Scales ggproto](https://ggplot2.tidyverse.org/reference/ggplot2-ggproto.html#scales):

> `transform()` Transforms a vector of values using self$trans. This occurs before the Stat is calculated.

Transformation changes the layer data

```{r 10-32}
toy # from Ch 10.1.5
ggplot(toy, aes(big, txt)) +
  geom_point()

reversed_plot <- ggplot(toy, aes(big, txt)) +
  geom_point() +
  scale_x_reverse()
reversed_plot
layer_data(reversed_plot)

rev_trans <- scales::reverse_trans()
scales::reverse_trans
str(rev_trans)
rev_trans$transform(toy$big)
rev_trans$inverse(rev_trans$transform(toy$big))
rev_trans$format(rev_trans$breaks(range(toy$big)))
```

Most useful for positioning purposes (ex: [`time_trans()`](https://scales.r-lib.org/reference/time_trans.html))

```{r 10-33}
hours <- seq(ISOdate(2000,3,20, tz = ""), by = "hour", length.out = 10)
t <- scales::time_trans()
t$transform(hours)
t$inverse(t$transform(hours))
t$format(t$breaks(range(hours)))
```

```{r 10-34}
date_trans_plot <- ggplot(tibble(hours = hours), aes(x = hours, y = 0)) +
  geom_point()
layer_data(date_trans_plot)
```


## 10.2 Date-time

### 10.2.1 Breaks

Book example:

```{r 10-35}
date_base <- ggplot(economics, aes(date, psavert)) + 
  geom_line(na.rm = TRUE) +
  labs(x = NULL, y = NULL)

date_base 
date_base + scale_x_date(date_breaks = "25 years")
```

Making it explicit:

```{r 10-36}
date_base + scale_x_date(breaks = scales::breaks_width("25 years"))
```

Book example:

```{r 10-37}
century20 <- as.Date(c("1900-01-01", "1999-12-31"))
breaks <- scales::breaks_width("25 years")
breaks(century20)
```

Using `offset` argument (unit = days):

```{r 10-38}
breaks2 <- scales::breaks_width("25 years", offset = 31) # offsets to Feb
breaks2(century20)
```

Calculating the offset:

```{r 10-39}
diff.Date(c(as.Date("1900-01-01"), as.Date("1900-02-01"))) # as.integer() to get value
```

### 10.2.2 Minor breaks

Book examples:

```{r 10-40}
date_base + scale_x_date(
  limits = as.Date(c("2003-01-01", "2003-04-01")),
  date_breaks = "1 month"
)

date_base + scale_x_date(
  limits = as.Date(c("2003-01-01", "2003-04-01")),
  date_breaks = "1 month",
  date_minor_breaks = "1 week"
)
```

> In the second plot, the major and minor beaks follow slightly different patterns: the minor breaks are always spaced 7 days apart but the major breaks are 1 month apart. Because the months vary in length, this leads to slightly uneven spacing.

Explicit:

```{r 10-41}
date_base + scale_x_date(
  limits = as.Date(c("2003-01-01", "2003-04-01")),
  breaks = scales::breaks_width("1 month")
)
date_base + scale_x_date(
  limits = as.Date(c("2003-01-01", "2003-04-01")),
  breaks = scales::breaks_width("1 month"),
  minor_breaks = scales::breaks_width("1 week")
)
```


### 10.2.3 Labels

Book examples:

```{r 10-42}
base <- ggplot(economics, aes(date, psavert)) + 
  geom_line(na.rm = TRUE) +
  labs(x = NULL, y = NULL)

base + scale_x_date(date_breaks = "5 years")
base + scale_x_date(date_breaks = "5 years", date_labels = "%y")
```

```{r 10-43}
base + scale_x_date(labels = scales::label_date_short())
lim <- as.Date(c("2004-01-01", "2005-01-01"))
base + scale_x_date(limits = lim, labels = scales::label_date_short())
```




## 10.3 Discrete

Book examples:

```{r 10-44}
ggplot(mpg, aes(x = hwy, y = class)) + 
  geom_point()

ggplot(mpg, aes(x = hwy, y = class)) + 
  geom_point() + 
  scale_x_continuous() +
  scale_y_discrete()

ggplot(mpg, aes(x = hwy, y = class)) + 
  geom_point() +
  annotate("text", x = 5, y = 1:7, label = 1:7)
```

### 10.3.1 Limits

> For discrete scales, limits should be a character vector that enumerates all possible values.

Censors missing categories in the set:

```{r 10-45}
ggplot(mpg, aes(x = hwy, y = class)) + 
  geom_point() +
  scale_y_discrete(limits = unique(mpg$class)[-1])
```

Adds new categories without value:

```{r 10-46}
ggplot(mpg, aes(x = hwy, y = class)) + 
  geom_point() +
  scale_y_discrete(limits = c("A", unique(mpg$class)))
```

Same effect with `drop = FALSE` with unused factor levels

```{r 10-47}
ggplot(mpg, aes(x = hwy, y = factor(class, levels = c("A", unique(class))))) +
  geom_point() +
  scale_y_discrete(drop = FALSE)
```

It drops unused factor levels by default, though

```{r 10-48}
ggplot(mpg, aes(x = hwy, y = factor(class, levels = c("A", unique(class))))) +
  geom_point() # + scale_y_discrete(drop = TRUE) 
```


### 10.3.2 Scale labels

```{r 10-49}
layer_data(last_plot()) %>% 
  ggplot(aes(x = x, y = y, group = group)) +
  geom_point()
```

### 10.3.2 Scale labels

Book example:

```{r 10-50}
base <- ggplot(toy, aes(const, txt)) + 
  geom_point() +
  labs(x = NULL, y = NULL)

base
base + scale_y_discrete(labels = c(c = "carrot", b = "banana"))
```

```{r 10-51}
base + scale_y_discrete(labels = str_to_title)
```

Debugging strategy

```{r 10-52, eval = FALSE}
browserer <- function(...) {
  params <- list(...)
  browser()
}
base + scale_y_discrete(labels = browserer)
```

### 10.3.3 `guide_axis()`

Book examples:

```{r 10-53}
base <- ggplot(mpg, aes(manufacturer, hwy)) + geom_boxplot() 

base + guides(x = guide_axis(n.dodge = 3))
base + guides(x = guide_axis(angle = 90))
```

More guides in `{ggh4x}` - [https://teunbrand.github.io/ggh4x/](https://teunbrand.github.io/ggh4x/index.html)

```{r 10-54}
library(ggh4x)
tibble(
  item = c("Coffee", "Tea", "Apple", "Pear", "Car"),
  type = c("Drink", "Drink", "Fruit", "Fruit", ""),
  amount = c(5, 1, 2, 3, 1)
) %>% 
  ggplot(aes(interaction(item, type), amount)) +
  geom_col() +
  scale_x_discrete(guide = guide_axis_nested()) # guides(x = "axis_nested")
```

## 10.4 Binned

Book example:

```{r 10-55}
base <- ggplot(mpg, aes(hwy, class)) + geom_count()

base 
base + scale_x_binned(n.breaks = 10)
```

```{r 10-56}
ggplot(mtcars, aes(hp)) +
  geom_histogram(binwidth = 20)

ggplot(mtcars, aes(hp)) +
  geom_bar() +
  scale_x_binned(breaks = scales::breaks_width(width = 20))
```

## ASIDE - `geom_sf()` + limits

### Example from Twitter:

[https://twitter.com/Josh_Ebner/status/1470818469801299970?s=20](https://twitter.com/Josh_Ebner/status/1470818469801299970?s=20)

### Reprexes from Ryan S:

```{r 10-57}
library(sf)
plygn_1 <- tibble(x_coord = c(1, 1, 2, 3, 6, 1),
                  y_coord = c(1, 2, 1, 2, 5, 1))
plygn_1
```

Full range polygon

```{r 10-58}
plygn_1 %>% 
  ggplot() +
  geom_polygon(aes(x_coord, y_coord))
```

Polygon with limits

```{r 10-59}
poly_plygn_1 <- plygn_1 %>%
  ggplot() +
  geom_polygon(aes(x_coord, y_coord)) +
  scale_x_continuous(limits = c(1, 4))
poly_plygn_1
```

Path with limits

```{r 10-60}
path_plygn_1 <- plygn_1 %>%
  ggplot() +
  geom_path(aes(x_coord, y_coord)) +
  scale_x_continuous(limits = c(1, 4))
path_plygn_1
```

`geom_sf()` without limits

```{r 10-61}
sf_plygn_1 <- plygn_1 %>%                 # tibble of coords
  as.matrix() %>%                         # make into a matrix
  list() %>%                              # make into a list
  st_polygon() %>%                        # make into sf object
  ggplot() +                              # call ggplot
  geom_sf()                             # use geom_sf for plotting sf objects

sf_plygn_1
```

`geom_sf()` with limits

```{r 10-62}
sf_plygn_1_wlims <- sf_plygn_1 +
  scale_x_continuous(limits = c(1, 4))    # add limits

sf_plygn_1_wlims
```

### Further exploration

Using `geom_sf()` adds `CoordSF` by default

```{r 10-63}
class(sf_plygn_1$coordinates)
class(sf_plygn_1_wlims$coordinates)
```

In fact, `geom_sf()` must be used with `coord_sf()`

```{r 10-64, error = TRUE}
# Same thing as without limits `sf_plygn_1`
sf_plygn_1 +
  coord_sf()

# Same thing as with limits `sf_plygn_1_wlims`
sf_plygn_1 +
  coord_sf(xlim = c(1, 4))

# Doesn't work
sf_plygn_1 +
  coord_cartesian()
```

The underlying geometry is untouched (indicating that limits are not removing data)

```{r 10-65}
layer_data(sf_plygn_1)
layer_data(sf_plygn_1_wlims)

identical(
  layer_data(sf_plygn_1_wlims)$geometry,
  layer_data(sf_plygn_1)$geometry
)
```

OOB handling inside `scale_x|y_continuous()` cannot override the behavior

```{r 10-66}
sf_plygn_1 +
  scale_x_continuous(limits = c(1, 4), oob = scales::oob_censor)
```

Instead, `coord_sf(lims_method = )` offers other spatial-specific methods. Censor doesn't seem to be one but an option like `"geometry_bbox"` automatically sets limits to the smallest bounding box that contain all geometries.

```{r 10-67}
sf_plygn_1 +
  coord_sf(lims_method = "geometry_bbox")
sf_plygn_1_wlims +
  coord_sf(lims_method = "geometry_bbox")
```

Interesting note from the [docs](https://ggplot2.tidyverse.org/reference/ggsf.html#arguments):

> ... specifying limits via position scales or xlim()/ylim() is strongly discouraged, as it can result in data points being dropped from the plot even though they would be visible in the final plot region.


### Internals

```{r 10-68}
library(ggtrace) # v.0.4.5
```

Scale censor for `geom_polygon()`

```{r 10-69, eval = FALSE}
ggbody(ggplot2:::ggplot_build.ggplot)[[17]]
ggtrace(
  method = ggplot2:::ggplot_build.ggplot,
  trace_steps = 17,
  trace_exprs = quote(browser()),
  verbose = FALSE
)
path_plygn_1
```

Scale censor for `geom_sf()`

```{r 10-70, eval = FALSE}
ggtrace(
  method = ggplot2:::ggplot_build.ggplot,
  trace_steps = 17, # and 26 `layout$map_position`
  trace_exprs = quote(browser()),
  verbose = FALSE
)
sf_plygn_1_wlims
```

Inspecting the rendered geom with `layer_grob()`

```{r 10-71}
patchwork::wrap_elements(layer_grob(poly_plygn_1)[[1]])
patchwork::wrap_elements(layer_grob(sf_plygn_1)[[1]])
patchwork::wrap_elements(layer_grob(sf_plygn_1_wlims)[[1]])

dplyr::bind_cols(layer_grob(sf_plygn_1)[[1]][c("x", "y")])
dplyr::bind_cols(layer_grob(sf_plygn_1_wlims)[[1]][c("x", "y")]) # x for fifth row is >1npc
```



## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/EvKchS3X6cg")`

`r knitr::include_url("https://www.youtube.com/embed/LSkVSJasHPY")`

<details>
  <summary> Meeting chat log </summary>
```
00:59:06	June Choe:	There's also a nice animation from wikipedia (the cylinder is squished because of perceptual inequality between hues) - https://upload.wikimedia.org/wikipedia/commons/transcoded/8/8d/SRGB_gamut_within_CIELCHuv_color_space_mesh.webm/SRGB_gamut_within_CIELCHuv_color_space_mesh.webm.480p.vp9.webm
```
</details>
