---
title: "11-Colour_scales_and_legends"
author: "Lydia Gibson"
date: "12/19/2021"
output: pdf_document
---


# Colour Scales and Legends

```{r library, include=FALSE}

library(ggplot2)

```


**Learning Objectives**

-Learn how to map values to colours in ggplot2

-Learn about colour theory (a more detailed exposition is available online at http://tinyurl.com/clrdtls)




## A little colour theory

There have been many attempts to come up with colours spaces that are more perceptually uniform. We’ll use a modern attempt called the HCL colour space, which has three components of hue, chroma and luminance:

-**Hue** ranges from 0 to 360 (an angle) and gives the “colour” of the colour (blue, red, orange, etc).

-**Chroma** is the “purity” of a colour, ranging from 0 (grey) to a maximum that varies with luminance.

-**Luminance** is the lightness of the colour, ranging from 0 (black) to 1 (white).

An additional complication is that many people (~10% of men) do not possess the normal complement of colour receptors and so can distinguish fewer colours than usual. In brief, it’s best to avoid red-green contrasts, and to check your plots with systems that simulate colour blindness. Visicheck (https://www.vischeck.com/vischeck/) is one online solution. Another alternative is the dichromat package34 which provides tools for simulating colour blindness, and a set of colour schemes known to work well for colour-blind people. You can also help people with colour blindness in the same way that you can help people with black-and-white printers: by providing redundant mappings to other aesthetics like size, line type or shape.

## Continuous colour scales

```{r continuous color scales 01}
erupt <- ggplot(faithfuld, aes(waiting, eruptions, fill = density)) +
  geom_raster() +
  scale_x_continuous(NULL, expand = c(0, 0)) + 
  scale_y_continuous(NULL, expand = c(0, 0)) + 
  theme(legend.position = "none")
```


### Particular pallettes


```{r particular pallettes 01}
erupt
erupt + scale_fill_viridis_c()
erupt + scale_fill_viridis_c(option = "magma")
```


```{r particular pallette 02}
erupt + scale_fill_distiller()
erupt + scale_fill_distiller(palette = "RdPu")
erupt + scale_fill_distiller(palette = "YlOrBr")

```




```{r particular pallettes 03}
erupt + scico::scale_fill_scico(palette = "bilbao") # the default
erupt + scico::scale_fill_scico(palette = "vik")
erupt + scico::scale_fill_scico(palette = "lajolla")
```


```{r particular pallettes 04}
erupt + paletteer::scale_fill_paletteer_c("viridis::plasma")
erupt + paletteer::scale_fill_paletteer_c("scico::tokyo")
erupt + paletteer::scale_fill_paletteer_c("gameofthrones::targaryen")
```




### Robust recipes



```{r robust recipes 01}
erupt + paletteer::scale_fill_paletteer_c("viridis::plasma")
erupt + paletteer::scale_fill_paletteer_c("scico::tokyo")
erupt + paletteer::scale_fill_paletteer_c("gameofthrones::targaryen")
```




```{r robust recipes 02}
erupt + paletteer::scale_fill_paletteer_c("viridis::plasma")
erupt + paletteer::scale_fill_paletteer_c("scico::tokyo")
erupt + paletteer::scale_fill_paletteer_c("gameofthrones::targaryen")
```



```{r robust recipes 03}
munsell::hue_slice("5P") +  # generate a ggplot with hue_slice()
  annotate(                 # add arrows for annotation 
    geom = "segment", 
    x = c(7, 7), 
    y = c(1, 10), 
    xend = c(7, 7), 
    yend = c(2, 9), 
    arrow = arrow(length = unit(2, "mm"))
  ) 
#> Warning: Removed 31 rows containing missing values (geom_text).

# construct scale
erupt + scale_fill_gradient(
  low = munsell::mnsl("5P 2/12"), 
  high = munsell::mnsl("5P 7/12")
)
```



```{r robust recipes 04}
# munsell example
erupt + scale_fill_gradient2(
  low = munsell::mnsl("5B 7/8"),
  high = munsell::mnsl("5Y 7/8"),
  mid = munsell::mnsl("N 7/0"),
  midpoint = .02
) 

# colorspace examples
erupt + scale_fill_gradientn(colours = colorspace::heat_hcl(7))
erupt + scale_fill_gradientn(colours = colorspace::diverge_hcl(7))
```


### Missing values



```{r missing values 01}
# munsell example
erupt + scale_fill_gradient2(
  low = munsell::mnsl("5B 7/8"),
  high = munsell::mnsl("5Y 7/8"),
  mid = munsell::mnsl("N 7/0"),
  midpoint = .02
) 

# colorspace examples
erupt + scale_fill_gradientn(colours = colorspace::heat_hcl(7))
erupt + scale_fill_gradientn(colours = colorspace::diverge_hcl(7))
```




### Limits, breaks and labels 


```{r limits break and labels 01}
# munsell example
erupt + scale_fill_gradient2(
  low = munsell::mnsl("5B 7/8"),
  high = munsell::mnsl("5Y 7/8"),
  mid = munsell::mnsl("N 7/0"),
  midpoint = .02
) 

# colorspace examples
erupt + scale_fill_gradientn(colours = colorspace::heat_hcl(7))
erupt + scale_fill_gradientn(colours = colorspace::diverge_hcl(7))
```


## Discrete colour scales

```{r discrete colour scales 01, include=FALSE}
df <- data.frame(x = c("a", "b", "c", "d"), y = c(3, 4, 1, 2))
bars <- ggplot(df, aes(x, y, fill = x)) + 
  geom_bar(stat = "identity") + 
  labs(x = NULL, y = NULL) +
  theme(legend.position = "none")
```

```{r discrete colour scales 02}
bars
bars + scale_fill_discrete()
bars + scale_fill_hue()
```




### Brewer scales


```{r brewer scales 01}
RColorBrewer::display.brewer.all()
```


```{r brewer scales 02}
bars + scale_fill_brewer(palette = "Set1")
bars + scale_fill_brewer(palette = "Set2")
bars + scale_fill_brewer(palette = "Accent")
```

```{r brewer scales 03}
# scatter plot
df <- data.frame(x = 1:3 + runif(30), y = runif(30), z = c("a", "b", "c"))
point <- ggplot(df, aes(x, y)) +
  geom_point(aes(colour = z))  + 
  theme(legend.position = "none") +
  labs(x = NULL, y = NULL)

# three palettes
point + scale_colour_brewer(palette = "Set1")
point + scale_colour_brewer(palette = "Set2")  
point + scale_colour_brewer(palette = "Pastel1")
```


```{r brewer scales 04}
# bar plot
df <- data.frame(x = 1:3, y = 3:1, z = c("a", "b", "c"))
area <- ggplot(df, aes(x, y)) + 
  geom_bar(aes(fill = z), stat = "identity") + 
  theme(legend.position = "none") +
  labs(x = NULL, y = NULL)

# three palettes
area + scale_fill_brewer(palette = "Set1")
area + scale_fill_brewer(palette = "Set2")
area + scale_fill_brewer(palette = "Pastel1")
```



### Hue and grey scales


```{r hue and gey scales 01}
bars
bars + scale_fill_hue(c = 40)
bars + scale_fill_hue(h = c(180, 300))
```


```{r hue and grey scales 02}
bars + scale_fill_grey()
bars + scale_fill_grey(start = 0.5, end = 1)
bars + scale_fill_grey(start = 0, end = 0.5)
```


### Manual scales


```{r manual scales 01}
bars + scale_fill_manual(values = c("sienna1", "sienna4", "hotpink1", "hotpink4"))
bars + scale_fill_manual(values = c("tomato1", "tomato2", "tomato3", "tomato4"))
bars + scale_fill_manual(values = c("grey", "black", "grey", "grey"))
```


```{r manual scales 02}
bars + scale_fill_manual(values = c(
  "d" = "grey",
  "c" = "grey",
  "b" = "black",
  "a" = "grey"
))
```


## Binned colour scales

```{r binned colour scales 01}
erupt + scale_fill_binned()
```


```{r binned colour scales 02}
erupt + scale_fill_steps()
```

```{r binned colour scales 03}
erupt + scale_fill_steps(n.breaks = 8)
```


```{r binned colour scales 04}
erupt + scale_fill_steps(low = "grey", high = "brown")
```

```{r binned colour scales 05}
erupt + scale_fill_steps2(low = "grey", mid = "white", high = "brown", midpoint = .02)
```


```{r binned colour scales 06}
erupt + scale_fill_stepsn(n.breaks = 12, colours = terrain.colors(12))
```



```{r binned colour scales 07}
erupt + scale_fill_fermenter(n.breaks = 9)
```


```{r binned colour scales 08}
erupt + scale_fill_fermenter(n.breaks = 9, palette = "Oranges")
```


```{r binned colour scales 09}
erupt + scale_fill_fermenter(n.breaks = 9, palette = "PuOr")
```


## Alpha scales


Alpha scales map the transparency of a shade to a value in the data. They are not often useful, but can be a convenient way to visually down-weight less important observations. scale_alpha() is an alias for scale_alpha_continuous() since that is the most common use of alpha, and it saves a bit of typing.





## Legends

### Positioning

```{r positioning 01, include=FALSE}
toy <- data.frame(
  const = 1, 
  up = 1:4,
  txt = letters[1:4], 
  big = (1:4)*1000,
  log = c(2, 5, 10, 2000)
)
toy
#>   const up txt  big  log
#> 1     1  1   a 1000    2
#> 2     1  2   b 2000    5
#> 3     1  3   c 3000   10
#> 4     1  4   d 4000 2000
axs <- ggplot(toy, aes(big, const)) + 
  geom_point() + 
  labs(x = NULL, y = NULL)
axs
axs + scale_x_continuous(breaks = scales::breaks_extended())
axs + scale_x_continuous(breaks = scales::breaks_extended(n = 2))
```



```{r positioning 02}
base <- ggplot(toy, aes(up, up)) + 
  geom_point(aes(colour = txt), size = 3) + 
  xlab(NULL) + 
  ylab(NULL)

base + theme(legend.position = "left")
base + theme(legend.position = "right") # the default 
base + theme(legend.position = "bottom")
base + theme(legend.position = "none")
```


```{r positioning 03}
base <- ggplot(toy, aes(up, up)) + 
  geom_point(aes(colour = txt), size = 3)

base + theme(legend.position = c(0, 1), legend.justification = c(0, 1))
base + theme(legend.position = c(0.5, 0.5), legend.justification = c(0.5, 0.5))
base + theme(legend.position = c(1, 0), legend.justification = c(1, 0))
```




## Legend key glyphs



```{r legend key glyphs 01}
base <- ggplot(economics, aes(date, psavert, color = "savings"))

base + geom_line()
base + geom_line(key_glyph = "timeseries")
```


### guide_legend()

```{r guide legend 01}
base <- ggplot(mpg, aes(drv, fill = factor(cyl))) + geom_bar() 

base
base + guides(fill = guide_legend(ncol = 2))
base + guides(fill = guide_legend(ncol = 2, byrow = TRUE))
```

```{r guide legend 02}
base
base + guides(fill = guide_legend(reverse = TRUE))
```



```{r guide legend 03}
base <- ggplot(mpg, aes(displ, hwy, colour = drv)) +
  geom_point(size = 4, alpha = .2, stroke = 0)

base + guides(colour = guide_legend())
base + guides(colour = guide_legend(override.aes = list(alpha = 1)))
```


### guide_bins()


```{r  guide bins 01, include=FALSE}
base <- ggplot(mpg, aes(displ, manufacturer, size = hwy)) +
  geom_point(alpha = .2) +
  scale_size_binned()
```


```{r guide bin 02}
base
base + guides(size = guide_bins(axis = FALSE))
```


```{r guide bins 03}
base + guides(size = guide_bins(direction = "vertical"))
base + guides(size = guide_bins(direction = "horizontal"))
```



```{r guide bins 04}
base + guides(size = guide_bins(show.limits = TRUE))
base + guides(
  size = guide_bins(
    axis.colour = "red",
    axis.arrow = arrow(
      length = unit(.1, "inches"), 
      ends = "first", 
      type = "closed"
    )
  )
)
```


### guide_colourbar()/guide_colorbar()


```{r guide colorbar 01}
base <- ggplot(mpg, aes(cyl, displ, colour = hwy)) +
  geom_point(size = 2)

base
base + guides(colour = guide_colourbar(reverse = TRUE))
base + guides(colour = guide_colourbar(barheight = unit(2, "cm")))
```


### 11.7.4 guidecoloursteps()/guidecolorsteps()

```{r guide colorstep 01}
base <- ggplot(mpg, aes(displ, hwy, colour = cyl)) + 
  geom_point() + 
  scale_color_binned()

base + guides(colour = guide_coloursteps(show.limits = TRUE))
base + guides(colour = guide_coloursteps(show.limits = FALSE))
```


