# Scales and Guides

```{r}
library(ggplot2)
```


**Learning objectives:**

* Show how concepts for position and scale apply elsewhere

* Theory underpinning scales and guides



## Theory of scales and guides

|Argument name|Axis| Label|
|-|-|-|
| `name` |Label| Title |
| `breaks` | Ticks & grid line|Key|
| `labels` |Tick label|Key label|

### Scale specification


```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(colour = class))
```

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(colour = class)) +
  scale_x_continuous() + 
  scale_y_continuous() + 
  scale_colour_discrete()
```

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(colour = class)) + 
  scale_x_continuous(name = "A really awesome x axis label") +
  scale_y_continuous(name = "An amazingly great y axis label")
```


```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  scale_x_continuous(name = "Label 1") +
  scale_x_continuous(name = "Label 2")
#> Scale for 'x' is already present. Adding another scale for 'x', which will
#> replace the existing scale.

ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  scale_x_continuous(name = "Label 2")
```

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(colour = class)) +
  scale_x_sqrt() + 
  scale_colour_brewer()
```



### Naming scheme





### Fundamental scale types


## Scale Names

## Scale Breaks

## Scale Limits

```{r}
df <- data.frame(x = 1:6, y = 8:13)
base <- ggplot(df, aes(x, y)) + 
  geom_col(aes(fill = x)) +                    # bar chart
  geom_vline(xintercept = 3.5, colour = "red") # for visual clarity only

base
base + scale_fill_gradient(limits = c(1, 3))
base + scale_fill_gradient(limits = c(1, 3), oob = scales::squish)
```



## Scale guides

|Scale type| Default guide type |
|-|-|
|continuous scales for colour / fill aesthetics | colourbar|
|binned scales for colour/fill aesthetics|coloursteps|
|position scales (continuous, binned, and discrete)| axis|
|discrete scales (except position scales)| legend|
|binned scalesd (except position/colour/fill scales)|bins|

## Scale transformation

```{r}
base <- ggplot(faithfuld, aes(waiting, eruptions)) + 
  geom_raster(aes(fill = density)) + 
  scale_x_continuous(NULL, NULL, expand = c(0, 0)) +
  scale_y_continuous(NULL, NULL, expand = c(0, 0))
  
base
base + scale_fill_continuous(trans = "sqrt")
```

```{r}
df <- data.frame(x = runif(20), y = runif(20), z = sample(20))
base <- ggplot(df, aes(x, y, size = z)) + geom_point()

base 
base + scale_size(trans = "reverse")
```


## Legend merging and splitting

### Merging legends

```{r}

```{r, echo=FALSE}
toy <- data.frame(
  const = 1, 
  up = 1:4,
  txt = letters[1:4], 
  big = (1:4)*1000,
  log = c(2, 5, 10, 2000)
)
```

```



```{r}
ggplot(toy, aes(up, up)) + 
  geom_point(size = 4, colour = "grey20") +
  geom_point(aes(colour = txt), size = 2) 

ggplot(toy, aes(up, up)) + 
  geom_point(size = 4, colour = "grey20", show.legend = TRUE) +
  geom_point(aes(colour = txt), size = 2) 
```



```{r}
base <- ggplot(toy, aes(const, up)) +
  scale_x_continuous(NULL, breaks = NULL)
base + geom_point(aes(colour = txt))
base + geom_point(aes(shape = txt))
base + geom_point(aes(shape = txt, colour = txt))
```


```{r}
base <- ggplot(toy, aes(const, up)) + 
  geom_point(aes(shape = txt, colour = txt)) + 
  scale_x_continuous(NULL, breaks = NULL)

base
base + labs(shape = "Split legend")
base + labs(shape = "Merged legend", colour = "Merged legend")
```


### Splitting legends


```{r}
base <- ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(colour = factor(year)), size = 5) + 
  scale_colour_brewer("year", type = "qual", palette = 5) 

base
base + 
  ggnewscale::new_scale_colour() + 
  geom_point(aes(colour = cyl == 4), size = 1, fill = NA) + 
  scale_colour_manual("4 cylinder", values = c("grey60", "black"))
```


## Legend key glyphs

```{r}
base <- ggplot(economics, aes(date, psavert, color = "savings"))

base + geom_line()
base + geom_line(key_glyph = "timeseries")
```




