# Coordinate systems


**Learning objectives:**


- What are `coord_<functions>` ?
- What are the differences between `coord_<functions>` in {ggplot2} ?
- How to use **coordinate systems** in {ggplot2}




## Introduction

The coordinate system in {ggplot2} can be managed with the use of `coord_<functions>`. This is done when we need to:

- zoom into a plot in a particular area of the plot
- flip the axis of a plot
- set a fixed aspect ratio of a plot

- transform coordinates
- change the shape of the plot
- set the coordinates for a map projection



```{r ch16-01, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE, warning=FALSE, paged.print=FALSE,comment = "")
```

```{r ch16-02}
library(tidyverse)
library(patchwork)
iris %>% head()
```



## Linear coordinate systems

- `coord_cartesian()`: the default Cartesian coordinate system, where the 2d position of an element is given by the combination of the x and y positions.

- `coord_flip()`: Cartesian coordinate system with x and y axes flipped.

- `coord_fixed()`: Cartesian coordinate system with a fixed aspect ratio.

----------


1.  `coord_cartesian()`

```{r ch16-03}
p1 <- ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
  geom_point(aes(fill=Species),
             show.legend = F,
             shape=21,color="grey20",alpha=0.5) + 
  geom_smooth(color="pink") +
  theme_light()

p1 | p1 + scale_x_continuous(limits = c(5, 6)) | p1 + coord_cartesian(xlim = c(5, 6))
```


2.  `coord_flip()`

```{r ch16-04}
p2 <- ggplot(iris, aes(Sepal.Length, Sepal.Width)) + 
  geom_point(aes(fill=Species),
             show.legend = F,
             shape=21,color="grey20",alpha=0.5) + 
  geom_smooth(color="pink") +
  theme_light()

p3 <- ggplot(iris, aes(Sepal.Width,Sepal.Length)) + 
  geom_point(aes(fill=Species),
             show.legend = F,
             shape=21,color="grey20",alpha=0.5) + 
  geom_smooth(color="pink") +
  theme_light()

p2 | p2 + coord_flip() | p3
```

(the smooth is fit to the rotated data).


3.  `coord_fixed()`

```{r ch16-05}
p3 | p3 + coord_fixed()
```





## Non-linear coordinate systems


- `coord_polar()`: Polar coordinates.

- `coord_map()`/`coord_quickmap()`/`coord_sf()`: Map projections.

- `coord_trans()`: Apply arbitrary transformations to x and y positions, after the data has been processed by the stat.


---------

1.  `coord_polar()`

```{r ch16-06}
p4 <- iris %>%
  ggplot(aes(x = Species, y = Petal.Width)) +
  geom_col(aes(color=Species,fill=Species),show.legend = F)+
  theme_light()


p4 + coord_polar(theta = "x") | p4 + coord_polar(theta = "y")
```


### Example: Coord_polar() with DuBoisChallenge N°8 data

source: [DuBois data portraits](https://github.com/ajstarks/dubois-data-portraits/tree/master/challenge/2022)

```{r ch16-07}
df <- read_csv("https://raw.githubusercontent.com/ajstarks/dubois-data-portraits/master/challenge/2022/challenge08/data.csv")

df2 <- df %>%
  arrange(-Year)  

df2[7,1]  <- 1875
df2[7,2]  <- 0
df2[7,3]  <- 0
```





```{r ch16-08,fig.align='center',fig.dim="100%"}
df2 %>%  
  ggplot() + 
  
  geom_line(data= subset(df2, Year %in% c(1875,1875)),
            mapping = aes(x=Year, y= `Houshold Value (Dollars)`),
            color="#FFCDCB",size=6) +
  
  geom_line(data= subset(df2, Year%in%c(1875,1875,1880)),
            mapping= aes(x=Year +2, y= `Houshold Value (Dollars)`),
            color="#989EB4",size=6) +
  
  geom_line(data= subset(df2, Year%in%c(1875,1875,1880,1885)),
            mapping= aes(x=Year +4, y= `Houshold Value (Dollars)`),
            color="#b08c71",size=6) +
  
  geom_line(data= subset(df2, Year%in%c(1875,1875,1880,1885,1890)),
            mapping= aes(x=Year +6, y= `Houshold Value (Dollars)`),
            color="#FFC942",size=6) +
  
  geom_line(data= subset(df2, Year%in%c(1875,1875,1880,1885,1890,1895)),
            mapping= aes(x=Year +8, y= `Houshold Value (Dollars)`),
            color="#EFDECC", size=6) +
  
  geom_line(mapping= aes(x=Year +10, y= `Houshold Value (Dollars)`),
            color="#F02C49",size=6) +
  
  coord_polar(theta = "y",
              start = 0,
              direction = 1, 
              clip = "off") +
  
  # other scales that can be used:
  #scale_x_reverse(expand=expansion(mult=c(-0.9,-0.1),add=c(29,-0.1))) +
  #scale_y_continuous(expand=expansion(mult=c(0.09,0.01),add=c(0,-790000))) +

  scale_x_reverse(expand=expansion(add=c(11,-5))) +
  scale_y_continuous(expand=expansion(add=c(0,-600000))) +
  labs(title="ASSESSED VALUE OF HOUSEHOLD AND KITCHEN FURNITURE 
       OWNED BY GEORGIA NEGROES.")+
  theme_void() +
  theme(text = element_text(face="bold",
                            color="grey27"),
        aspect.ratio =2/1.9, #y/x
        plot.background = element_rect(color= "#d9ccbf", fill= "#d9ccbf"),
        plot.title = element_text(hjust=0.5,size=9))
```

2.  `coord_trans()`

```{r ch16-09}
rect <- data.frame(x = 50, y = 50)
line <- data.frame(x = c(1, 200), y = c(100, 1))
p6 <- ggplot(mapping = aes(x, y)) + 
  geom_tile(data = rect, aes(width = 50, height = 50)) + 
  geom_line(data = line) + 
  xlab(NULL) + ylab(NULL)

p6
```

```{r ch16-10}
p6 + coord_trans(y = "log10") 
```


```{r ch16-11}
p7 <- ggplot(iris, aes(Sepal.Length, Petal.Length)) + 
  stat_bin2d() + 
  geom_smooth(method = "lm") + 
  xlab(NULL) + 
  ylab(NULL) + 
  theme(legend.position = "none")
p7
#> `geom_smooth()` using formula 'y ~ x'

# Better fit on log scale, but harder to interpret
p7 +
  scale_x_log10() + 
  scale_y_log10()
#> `geom_smooth()` using formula 'y ~ x'

# Fit on log scale, then backtransform to original.
# Highlights lack of expensive diamonds with large carats
pow10 <- scales::exp_trans(10)
p7 +
  scale_x_log10() + 
  scale_y_log10() + 
  coord_trans(x = pow10, y = pow10)
```


3.  `coord_map()`/`coord_quickmap()`/`coord_sf()`

```{r ch16-12}
world <- map_data("world")
worldmap <- ggplot(world, aes(long, lat, group = group)) +
  geom_path() +
  scale_y_continuous(NULL, breaks = (-2:3) * 30, labels = NULL) +
  scale_x_continuous(NULL, breaks = (-4:4) * 45, labels = NULL)


worldmap + coord_quickmap() |
worldmap + coord_map("ortho") |
worldmap + coord_map("stereographic")
```


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary> Meeting chat log </summary>
  
```
```
</details>
