# (PART\*) Advanced Topics {-}

# Programming with ggplot2

**Learning objectives:**

- Programming single and multiple components
- Use components, annotation, and additional arguments in a plot
- Functional programming

---

```{r message=FALSE, warning=FALSE, include=FALSE, paged.print=FALSE}
library(tidyverse)
```

## Why program with {ggplot2}? {-}

To reduce duplicated code, build up repeated components

Can also generalize code to allow for flexibility

### Plot components are objects! {-}

One example of a component of a plot is this one below:
```{r}
bestfit <- geom_smooth(
  method = "lm", 
  se = FALSE, 
  colour = alpha("steelblue", 0.5), 
  linewidth = 2)
class(bestfit)
```

## Adding the object we created to a plot {-}

Place component in grammar of graphics syntax to use as a plot layer

```{r}
ggplot(mpg, aes(cty, hwy)) + 
  geom_point() + 
  bestfit
```

## Creating a function {-}

Another way is to build a layer via a function:
```{r}
geom_lm <- function(formula = y ~ x, colour = alpha("steelblue", 0.5), 
                    linewidth = 2, ...)  {
  geom_smooth(formula = formula, se = FALSE, method = "lm", colour = colour,
    linewidth = linewidth, ...)
}
```

Use the layer in the plot:
```{r}
ggplot(mpg, aes(displ, 1 / hwy)) + 
  geom_point() + 
  geom_lm(y ~ poly(x, 2), linewidth = 1, colour = "red")
```

## `...` gives functions flexibility {-}

The `...` parameter lets a function accept additional arbitrary arguments

e.g. `na.rm` (doesn't make a difference in this case, but the function accepts it!)

```{r}
ggplot(mpg, aes(displ, 1 / hwy)) + 
  geom_point() + 
  geom_lm(y ~ poly(x, 2), linewidth = 1, colour = "red", na.rm = T)
```

## Exercises {-}

1. Create an object that represents a pink histogram with 100 bins.
```{r}
pinkhist <- geom_histogram(fill = "pink", bins = 100)
ggplot(diamonds, aes(x = price)) +
  pinkhist +
  labs(y = "Frequency",
       x = "Price")
```
2. Create an object that represents a fill scale with the Blues ColorBrewer palette.
```{r}
blues_fill_scale <- scale_fill_brewer(palette = "Blues")
ggplot(data = diamonds, aes(x = cut, y = price, fill = cut))+
  geom_boxplot()+
  theme_minimal()+
  blues_fill_scale
```

3. Read the source code for theme_grey(). What are its arguments? How does it work?
```{r}
theme_grey
# It creates a theme object called t
# uses %+replace% to replace the existing theme with t

# ggplot_global$theme_all_null %+replace% t
# ggplot_global$theme_all_null doesn't exist globally, so must refer to the current plot that you're adding theme_grey to.
```

4. Create scale_colour_wesanderson(). It should have a parameter to pick the palette from the wesanderson package, and create either a continuous or discrete scale.
```{r}
library(wesanderson)
scale_colour_wesanderson <- function(palette, type = "discrete", ...){
  scale_color_manual(values = wes_palette(name = palette, type = type, ...), ...)
}
ggplot(diamonds, aes(x = carat, y = price, color = cut))+
  geom_point()+
  scale_colour_wesanderson(palette = "Cavalcanti1")+
  theme_minimal()
```

## A ggplot object is a list!

And therefore, we can add more than one component as a list.

```{r}
geom_mean <- function() {
  list(
    stat_summary(fun = "mean", geom = "bar", fill = "grey70"),
    stat_summary(fun.data = "mean_cl_normal", geom = "errorbar", width = 0.4)
  )
}
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
ggplot(mpg, aes(class, cty)) + geom_mean()
```


## Components of a plot {-}

- data.frame
- aes()
- Scales
- Coords systems
- Theme components

## We can add any of these to a plot and override existing {-}

For datasets, use `%+%`

```{r}
dataset1 <- data.frame(id = rep(letters[1:3], each = 100),
                       Value = c(rnorm(100, mean = 1),
                                 rnorm(100, mean = 2),
                                 rnorm(100, mean = 3)))

dataset2 <- data.frame(id = rep(letters[1:3], each = 100),
                       Value = c(rpois(100, lambda = 1),
                                 rpois(100, lambda = 2),
                                 rpois(100, lambda = 3)))
p1 <- ggplot(dataset1, aes(x = Value, col = id))+
  geom_density()+
  theme_minimal()
p1

p2 <- p1 %+% dataset2
p2
```

## What if the dataset doesn't have the same variables? {-}

```{r error = T}
dataset3 <- data.frame(id = rep(letters[1:3], each = 100),
                       test = c(rpois(100, lambda = 4),
                                 rpois(100, lambda = 5),
                                 rpois(100, lambda = 6)))

# Try to add a new dataset, but it doesn't work because the code for p1 is expecting a "Value" column and that column doesn't exist in dataset3.
p1 %+%
  dataset3
```

Why doesn't this work?

```{r}
# Let's override the y aesthetic...
new_aes <- aes(y = test)

p3 <- p1 +
  new_aes %+%
  dataset3 # 
p3
```

## Annotations {-}

Make sure to set `inherit.aes = FALSE` and `show.legend = FALSE`

> "A quick and dirty way to get map data (from the maps package) on to your plot."

```{r}
borders <- function(database = "world", regions = ".", fill = NA, 
                    colour = "grey50", ...) {
  df <- map_data(database, regions)
  geom_polygon(
    aes_(~long, ~lat, group = ~group), 
    data = df, fill = fill, colour = colour, ..., 
    inherit.aes = FALSE, show.legend = FALSE
  )
}
```

```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(maps)
data(us.cities)
capitals <- subset(us.cities, capital == 2)

ggplot(capitals, aes(long, lat)) +
  borders("world", xlim = c(-130, -60), ylim = c(20, 50)) +
  geom_point(aes(linewidth = pop)) +
  scale_size_area() +
  coord_quickmap()
```

## Additional arguments {-}

"If you want to pass additional arguments to the components in your function, ... is no good: there’s no way to direct different arguments to different components. Instead, you’ll need to think about how you want your function to work, balancing the benefits of having one function that does it all vs. the cost of having a complex function that’s harder to understand."

      modifyList()
      do.call()

```{r}
geom_mean <- function(..., bar.params = list(), errorbar.params = list()) {
  params <- list(...)
  bar.params <- modifyList(params, bar.params)
  errorbar.params  <- modifyList(params, errorbar.params)
  
  bar <- do.call("stat_summary", modifyList(
    list(fun = "mean", geom = "bar", fill = "grey70"),
    bar.params)
  )
  errorbar <- do.call("stat_summary", modifyList(
    list(fun.data = "mean_cl_normal", geom = "errorbar", width = 0.4),
    errorbar.params)
  )

  list(bar, errorbar)
}
```
      
And here is the result:
```{r}
ggplot(mpg, aes(class, cty)) + 
  geom_mean(
    colour = "steelblue",
    errorbar.params = list(width = 0.5, size = 1)
  )
```

## Making the complete plot--very limited flexibility {-}
```{r}
piechart <- function(data, mapping) {
  ggplot(data, mapping) +
    geom_bar(width = 1) + 
    coord_polar(theta = "y") + 
    xlab(NULL) + 
    ylab(NULL)
}
piechart(mpg, aes(factor(1), fill = class))
```

## What if we want to pass in different variables? {-}

- Instead of writing out the entire `aes()`, the user can just pass in the variable names
- But there's a catch!

This doesn't work:
```{r}
my_function <- function(x_var) {
  aes(x = x_var)
}
my_function(abc)
#> Aesthetic mapping: 
#> * `x` -> `x_var`
```

We can "embrace" the argument to tell ggplot2 to "look inside" the argument and use its value, not its expression
```{r}
my_function <- function(x_var) {
  aes(x = {{x_var}})
}
my_function(abc)
#> Aesthetic mapping: 
#> * `x` -> `abc`
```

New version of the piechart function:
```{r}
piechart <- function(data, var) {
  ggplot(data, aes(factor(1), fill = {{ var }})) +
    geom_bar(width = 1) + 
    coord_polar(theta = "y") + 
    xlab(NULL) + 
    ylab(NULL)
}
mpg |> piechart(class)
```


## References {-}

- [extending ggplot2](https://ggplot2.tidyverse.org/articles/extending-ggplot2.html)
- [functions](https://adv-r.hadley.nz/functions.html)
- [expressions](http://adv-r.had.co.nz/Expressions.html)
- [functional programming](http://adv-r.had.co.nz/Functional-programming.html)
- [advanced R - functionals](https://adv-r.hadley.nz/fp.html)

---

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/jf-Qn4iFqHY")`

<details>
  <summary> Meeting chat log </summary>
  
```
00:41:31	Priyanka Gagneja:	There’s a lot of disturbance :(
01:00:48	Priyanka Gagneja:	https://plotly.com/ggplot2/setting-graph-size/
```
</details>
