# Themes

**Learning objectives:**


- How can I customize the output of my plot
- What are the functions `theme_<function>()` and `theme()`

## Theme


Plots can be customized by adding these function to your plot:

- scale_fill/color_<function>
- theme_<function>
- theme()
- ...

### Complete themes

In ggplo2 there are preset themes ready to use:
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
library(tidyverse)
df <- data.frame(x = 1:3, y = 1:3)
base <- ggplot(df, aes(x, y)) + geom_point()
p1<-base + theme_grey() + ggtitle("theme_grey()")
p2<-base + theme_bw() + ggtitle("theme_bw()")
p3<-base + theme_linedraw() + ggtitle("theme_linedraw()")

library(patchwork)
p1+p2+p3
```

```{r}
p4<-base + theme_light() + ggtitle("theme_light()")
p5<- base + theme_dark() + ggtitle("theme_dark()")
p6<-base + theme_minimal()  + ggtitle("theme_minimal()")

p4+p5+p6
```


```{r}
p7<-base + theme_classic() + ggtitle("theme_classic()")
p8<-base + theme_void() + ggtitle("theme_void()")

p7+p8
```

Or, you can use other packages such as {ggthemes} or other here: [ggplot extension gallery](https://exts.ggplot2.tidyverse.org/gallery/)

```{r}
library(ggthemes)
p9<-base + theme_tufte() + ggtitle("theme_tufte()")
p10<-base + theme_solarized() + ggtitle("theme_solarized()")
p11<-base + theme_excel() + ggtitle("theme_excel()") 

p9+p10+p11
```

- Modifying complete theme components with `theme()` function

## Plot elements of a theme

- Axis elements
- Legend elements
- Panel elements
- Faceting elements

Look at `?theme()` funtion in your help pane of RStudio for more info.

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>
  <summary> Meeting chat log </summary>
  
```
```
</details>
