--- 
title: "ggplot2 Book Club"
author: "The R4DS Online Learning Community"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: book.bib
biblio-style: apalike
link-citations: yes
github-repo: r4ds/bookclub-ggplot2
description: "This is the product of the R4DS Online Learning Community's Book Club."
---
--- 
title: "ggplot2 Book Club"
author: "The R4DS Online Learning Community"
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: book.bib
biblio-style: apalike
link-citations: yes
github-repo: r4ds/bookclub-ggplot2
description: "This is the product of the R4DS Online Learning Community's Book Club."
---

# Welcome {-}

This is a companion for the book [ggplot2: Elegant Graphics for Data Analysis](https://ggplot2-book.org/index.html) by Hadley Wickham, Danielle Navarro, and Thomas Lin Pedersen.
This companion is available at [https://r4ds.github.io/bookclub-ggplot2](https://r4ds.github.io/bookclub-ggplot2).

This website is being developed by the [R4DS Online Learning Community](https://rfordatasci.com). Follow along, and [join the community](https://r4ds.io/join) to participate.

This companion follows the [R4DS Online Learning Community Code of Conduct](https://r4ds.io/conduct).

## Book club meetings {-}

- Each week, a volunteer will present a chapter from the book.
  - **This is the best way to learn the material.**
- Presentations will usually consist of a review of the material, a discussion, and/or a demonstration of the principles presented in that chapter.
- More information about how to present is available in the [github repo](https://github.com/r4ds/bookclub-ggplot2).
- Presentations will be recorded, and will be available on the [R4DS Online Learning Community YouTube Channel](https://r4ds.io/youtube).

<!--chapter:end:index.Rmd-->

# Introduction

**Learning objectives:**

- {These are nice to have, but take some extra work. It's ok to skip these if necessary.}

## {Slide 1 title}

- {Create slides as sections, but keep them shrot like a slide.}

## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/jrBiEppKt_0")`

<details>
  <summary> Meeting chat log </summary>
  
```
```
</details>

<!--chapter:end:01-introduction.Rmd-->

---
title: "02-First_steps"
output: ioslides_presentation
---

# Update to Slides required

<!--chapter:end:02-First_steps.Rmd-->

---
title: "03-Individual_geoms"
output: ioslides_presentation
---

# Update to Slides required

<!--chapter:end:03-Individual_geoms.Rmd-->

---
title: "04-Collective_geoms"
output: ioslides_presentation
---

# Update to Slides required

<!--chapter:end:04-Collective_geoms.Rmd-->


# Statistical Summaries

Placeholder


## Defintions (in this Chapter)
## Revealing Uncertainty
## Weighted Data
## Displaying distributions
### Exercise:
## Dealing with overplotting
## Statistical Summaries

<!--chapter:end:05-Statistical_summaries.Rmd-->

---
title: "06-Maps"
author: "Ryan L. Metcalf"
date: "10/18/2021"
output: html_document
---
# Maps

**Learning Objectives:**
 - Plot simple maps using `geom_polygon()`
 - Using *simple features* `sf` to plot GIS data `geom_sf()`
 - Work with map projections and underlying `sf` data structure
 - Draw maps using Raster data
 
Plotting geospacial data is a common visualization task. The process may require spcialized tools. You can decompse the problem into two paths:
  - Using one data source to draw a map (if you have GIS data)
  - Adding metadata from another information source to the map (more common with relation to geographic areas)
  
> NOTE: X = Longitude, Y=Latitude. When pronounced "Lat/Lon" it is actually measured as Y/X. Not confusing....just keeping with vocabulary and measurements!

## Polygon Maps
The simplest approach to mapping is using geom_polygon(). This forms bounderies around regions.

```{r mi-countries, include=TRUE}
library(ggplot2)
mi_counties <- map_data("county", "michigan") %>% 
  select(lon = long, lat, group, id = subregion)
head(mi_counties)
```

In this data set we have four variables:
 - `lat`: Latitude of the vertex (as measured by horizontal paths)
 - `long`: Longitude of the vertex (as measured by vertical paths)
 - `id`: name of the region
 - `group`: unique identifier for contiguous areas within a region
 
```{r coord_quickmap(), include=TRUE}
ggplot(mi_counties, aes(lon, lat)) + 
  geom_point(size = .25, show.legend = FALSE) +
  coord_quickmap()
ggplot(mi_counties, aes(lon, lat, group = group)) +
  geom_polygon(fill = "white", colour = "grey50") + 
  coord_quickmap()
```

In this plot, `coord_quickmap()` is used to adjust the axes to ensure longitude and latitude are rendered on the same scale

For a more advanced use of ggplot2 for mapping, we'll see the use of `geom_sf()` and `coord_sf()` to handle spatial data specified in simple features format.

## Simple Features Maps
You *can* use the above examples...but not real world pratical. Instead, most GIS data is written as *simple features* and produced by the (Open Geospatial Consortium]https://www.ogc.org/)

### Layered Maps
### Labelled Maps
### Adding Other Geoms
## Map Projections
## Working with `sf` Data
## Raster Maps
## Data Sources


<!--chapter:end:06-Maps.Rmd-->

